---
title: "Fish BLT - Data Pipeline"
date: today
format: 
  html:
    theme: minty
    self-contained: true
    code-fold: true
    toc: true 
    toc-depth: 3
    toc-location: right
    html-table-processing: none
execute:
  fig-width: 10
---

```{r setup, message = F, warning = F, fig.width = 10, fig.height = 10, echo = F}
options(scipen = 999)

# Hook to format inline numeric expressions with comma separators:
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {return(as.character(x))}
  format(x, big.mark = ",", scientific = FALSE)})

library(sf)
library(hms)
library(readxl)
library(janitor)
library(lubridate)
library(gt)
library(pointblank)
library(tidyverse)
library(bigrquery)
library(leaflet)
library(leaflet.extras)
library(ggtext)
library(PristineSeasR2)

# ---- Parameters -----------------------------------------------------------
exp_id   <- "CHL_2020"

ps_paths <- PristineSeasR2::get_drive_paths()

prj_path <- file.path(ps_paths$projects, "legacy-db")

# ---- Auth & Connection ----------------------------------------------------
bigrquery::bq_auth(email = "marine.data.science@ngs.org")
bq_connection <- DBI::dbConnect(bigrquery::bigquery(), project = "pristine-seas")

set_theme(PristineSeasR2::theme_pristineseas())

data_out <- file.path(prj_path, "data/output/uvs")

figures_out <- file.path(prj_path, "figures/uvs")

legacy_path <- file.path(ps_paths$science, "legacy data/Kawesqar_Chilean_Fjords/")
```

# Overview

This documentation outlines the end-to-end pipeline for processing fish Belt Line Transect (BLT) survey data collected during Pristine Seas expeditions. The pipeline ingests raw field data, standardizes formats, performs taxonomy lookups, applies rigorous quality assurance/quality control (QA/QC), computes station-level summaries, and prepares validated outputs for the Pristine Seas Science Database in BigQuery.

# 1. Data Ingestion

## Reference Tables

We begin by reading the UVS (Underwater Visual Survey) site information already processed.

```{r load_references}
uvs_sites <- tbl(bq_connection, "uvs.sites") |> 
   filter(exp_id == !!exp_id) |> 
   collect()

# Validate site data
cat("Sites loaded:", nrow(uvs_sites), "\n")
cat("Regions:",      unique(uvs_sites$region), "\n")
cat("Subregions:",   unique(uvs_sites$subregion), "\n")

# Fish taxonomy reference
fish_taxa_lut <- tbl(bq_connection, "taxa_info.fish") |> 
  filter(!is.na(accepted_name)) |> 
  collect()

cat("✓ Loaded", nrow(fish_taxa_lut), "fish taxa\n")
```

## Fieldbooks

```{r}
raw_fish_obs <- read_xlsx(file.path(legacy_path, "Fish/Metadata_Fish_Kaweskar_V4.xlsx")) |> 
  clean_names() |> 
  select(diver, site = station, transect, depth_m, taxon_code = species, count = number, avg_tl_cm = avg)

taxa_codes <- read_xlsx(file.path(legacy_path, "Fish/Kewasqar_fish_codes_MH.xlsx")) |> 
  clean_names() |> 
  distinct(taxon_code = taxa_codes, species_name, trophic_group = trophic, lw_a = a, lw_b = b, lw_source = l_w_ref)

taxa_codes$species_name[taxa_codes$taxon_code == "Austrolicus depressiceps"] <- "Austrolycus depressiceps"
```

# 2. Data Transformation

```{r}
fish_obs <- raw_fish_obs |> 
  mutate(exp_id = exp_id,
         site = str_replace(site, "(\\d+)$", ~ str_pad(.x, 3, pad = "0")),
         ps_site_id = paste(exp_id, "uvs", str_pad(site, 3, pad = "0"), sep = "_"),
         transect_length_m = 25,
         transect_width    = 2,
         obs_id            = paste(exp_id,"fish", diver, str_pad(row_number(), 3, pad = "0"), sep = "_")) |> 
  select(obs_id, exp_id, ps_site_id, diver, depth_m, transect, transect_length_m, transect_width, taxon_code, length_cm = avg_tl_cm,  count)

transect_depths <- read_xlsx(file.path(legacy_path, "Benthos/Cape_Horn_Kawesqar_kelp_density.xlsx"), sheet = "Kawesqar") |> 
  janitor::clean_names() |> 
  mutate(depth_m           = round(depth),
         transect          = LETTERS[transect],
         ps_site_id        = paste(exp_id, "uvs", str_pad(station, 3, pad = "0"), sep = "_")) |> 
  distinct(ps_site_id, transect, depth_m)

fish_obs <- fish_obs |> 
  select(-depth_m) |> 
  left_join(transect_depths, by = c("ps_site_id", "transect"))

fish_obs$depth_m[fish_obs$ps_site_id == "CHL_2020_uvs_001"] <- 11
```

```{r}
# --- Derive transect- and station-level metadata ------------------------------
# Each transect represents a distinct belt within a site. Transect area incorporates both small-fish (2 m) and large-fish (4 m) belts. 

## Aggregate transects ---------------------------------------------------------
transects <- fish_obs |> 
  group_by(ps_site_id, diver, transect, transect_length_m, transect_width) |> 
  summarise(avg_depth_m = round(mean(depth_m, na.rm = TRUE), 0),
            .groups = "drop") |> 
  mutate(transect_area_m2 = transect_length_m * transect_width)  

## Aggregate stations ----------------------------------------------------------

stations <- transects |> 
  group_by(ps_site_id,  diver) |>
  summarise(n_transects     = n(),
            avg_depth_m     = round(mean(avg_depth_m, na.rm = TRUE), 0),
            survey_dist_m = sum(transect_length_m, na.rm = TRUE),
            survey_area_m2 = sum(transect_area_m2, na.rm = TRUE),
            .groups = "drop") |> 
  mutate(depth_strata = stratify(avg_depth_m),
         ps_station_id = paste0(ps_site_id, "_", station_suffix(depth_strata))) |> 
  select(ps_station_id, ps_site_id, depth_m = avg_depth_m, 
         depth_strata, divers = diver, n_transects, survey_dist_m, survey_area_m2)

# Assign station IDs and strata

fish_obs <- fish_obs |> 
  left_join(stations |> 
              distinct(ps_station_id, ps_site_id, depth_strata),
            by = c("ps_site_id")) |> 
  relocate(ps_station_id, .after = ps_site_id) |> 
  relocate(depth_strata, .after = ps_station_id)
```

# 3. Taxonomy Resolution

Match observed species with the reference taxonomy database and identify any issues (observed lengths > max lengths).

```{r}
fish_taxa <- fish_obs |> 
  distinct(taxon_code) |> 
  left_join(taxa_codes |> 
              select(taxon_code, taxon = species_name)) |> 
  mutate(taxon_clean = bdc::bdc_clean_names(taxon)$names_clean)

clean_names <- fish_taxa$taxon_clean |> 
  unique() |> 
  na.omit()  # drops NA

batches <- split(clean_names, ceiling(seq_along(clean_names) / 100))

# === 3. Query WoRMS in batches

worms_raw <- purrr::map_dfr(batches, ~worrms::wm_records_names(.x)) |> 
  select(taxon_clean = scientificname, 
         rank,
         aphia_id = AphiaID, 
         name_status = status, 
         accepted_name = valid_name, 
         accepted_aphia_id = valid_AphiaID) 

get_taxonomic_ranks <- function(id) {
  tryCatch(
    {worrms::wm_classification(id) |>
        select(rank, scientificname) |>
        pivot_wider(names_from = rank, values_from = scientificname) |>
        mutate(accepted_aphia_id = id)
      },
    error = function(e) {
      tibble(accepted_aphia_id = id)
    })}

library(furrr)

taxonomy_ranks <- future_map_dfr(worms_raw$accepted_aphia_id,
                                 get_taxonomic_ranks,
                                 .options = furrr_options(seed = TRUE)) |>
  clean_names() |>
  select(accepted_aphia_id, kingdom, phylum, class, order, family, genus)

worms_final <- worms_raw |> 
  left_join(distinct(taxonomy_ranks), by = "accepted_aphia_id")

fish_taxa <- fish_taxa |> 
  left_join(worms_final)

fish_taxa <- fish_taxa |> 
  left_join(taxa_codes |> 
              select(trophic_group, taxon_code, lw_a, lw_b))

fish_taxa$trophic_group <- "lower_carnivore"
```

## Join Taxonomy to Observations

```{r}
fish_obs <- fish_obs |> 
  left_join(fish_taxa, by = c("taxon_code")) |> 
  select(obs_id, ps_station_id, exp_id,
         diver, depth_m, transect, transect_length_m, transect_width_m = transect_width,
         accepted_name, accepted_aphia_id, rank, family, trophic_group,
         count, length_cm, lw_a, lw_b)
```

# 4.  QA/QC Validation

### Observations

```{r}
# Observation QA/QC

create_agent(tbl = fish_obs, 
             label = "Fish BLT Observations QA/QC", 
             tbl_name = "raw_obs") |> 
  col_vals_not_null(columns = vars(ps_station_id, transect, accepted_name, count),
                    label = "Core fields are not null",
                    actions = action_levels(warn_at = 0.001)) |>
  col_vals_gte(columns = vars(count),
               value = 1,
               label = "Count is positive",
               actions = action_levels(stop_at = 0.001)) |>
  interrogate()
```

### Trasects

```{r}
transects |> 
  create_agent(label = "Fish BLT Tranects QA/QC", tbl_name = "transects") |> 
  # Core fields not null
  col_vals_not_null(columns = vars(ps_site_id, diver, transect),
                    label = "Core fields are not null",
                    actions = action_levels(warn_at = 0.001)) |> 
  # Depth is less than 30 
  col_vals_lte(columns = vars(avg_depth_m),
               value   = 30,
               label = "Depth is less than 30 m",
               actions = action_levels(warn_at = 0.001)) |>
  # Transect length is 25
  col_vals_equal(columns = vars(transect_length_m), 
                 value = 25,
                 label = "Standard Transect length",
                 actions = action_levels(warn_at = 0.001)) |>
  interrogate()
```


### Stations

```{r}
stations |> 
  create_agent(label = "Fish BLT Stations QA/QC", tbl_name = "raw_stations") |> 
  # Core fields not null
  col_vals_not_null(columns = vars(ps_station_id, ps_site_id, depth_m, depth_strata, n_transects),
                    label = "Core fields are not null",
                    actions = action_levels(warn_at = 0.001)) |>
  # Unique station IDs
  rows_distinct(ps_station_id,
                label = "Station IDs are unique",
                actions = action_levels(warn_at = 0.001)) |> 
  col_vals_equal(columns = vars(n_transects), 
                 value = 2,
                 label = "2 transects per station",
                 actions = action_levels(warn_at = 0.001)) |>
  interrogate()
```

```{r}
# Create final station dataframe and join with site metadata

stations <- stations |> 
  left_join(uvs_sites |> 
              select(ps_site_id, region, exp_id, subregion, locality, habitat, exposure),
            by = "ps_site_id") |> 
  select(ps_station_id, ps_site_id, exp_id,
         region, subregion, locality, habitat, exposure,
         depth_strata, depth_m, divers,
         n_transects, survey_dist_m, survey_area_m2) 

stations$divers <- "Mathias Hune"
```

# 5. Calculate Biomass

```{r}
# Per-observation biomass

biomass_per_obs <- fish_obs |> 
  mutate(ltl_ratio = 1,
         # Biomass calculations
         avg_biomass   = lw_a * (length_cm * ltl_ratio)^lw_b,
         biomass_g     = round(count * avg_biomass, 4), # round to 4 digits
         ) |> 
  select(-lw_a, -lw_b, -ltl_ratio, -avg_biomass)

# Transect x taxa aggregation

biomass_by_transect_and_taxa <- biomass_per_obs |> 
  mutate(# Size-dependent transect area
         density_m2    = count / (transect_length_m*transect_width_m),
         biomass_gm2   = round(biomass_g / (transect_length_m*transect_width_m), 4)) |> 
  group_by(ps_station_id, diver, transect, 
           accepted_name, accepted_aphia_id, rank, family, trophic_group) |> 
  summarize(sum_length_count = sum(length_cm * count, na.rm = TRUE),
            min_length = min(length_cm, na.rm = TRUE),
            max_length = max(length_cm, na.rm = TRUE),
            across(c(count, biomass_g, biomass_gm2, density_m2), 
                   \(x) sum(x, na.rm = TRUE)),
            .groups = "drop")

# Complete missing taxa-transect combos within region

biomass_by_station_and_taxa <- biomass_by_transect_and_taxa |>
  left_join(stations |> 
              distinct(ps_station_id, region),
            by = "ps_station_id") |>
  group_by(region) |> 
  complete(nesting(ps_station_id, diver, transect),
           nesting(accepted_name, accepted_aphia_id, rank, family, trophic_group),
           fill = list(sum_length_count = NA,
                       min_length       = NA,
                       max_length       = NA,
                       count            = 0,
                       biomass_g        = 0,
                       density_m2       = 0,
                       biomass_gm2      = 0)) |> 
  group_by(ps_station_id, accepted_name, accepted_aphia_id, rank, family, trophic_group) |>
  summarise(total_count       = sum(count, na.rm = TRUE),
            avg_count         = mean(count),
            avg_length_cm     = round(if_else(total_count > 0, sum(sum_length_count, na.rm = TRUE) / sum(total_count, na.rm = TRUE), NA_real_),2),
            min_length_cm     = if_else(total_count > 0, 
                                        if (all(is.na(min_length))) NA_real_ else min(min_length, na.rm = TRUE), 
                                        NA_real_),
            max_length_cm     = if_else(total_count > 0, 
                                        if (all(is.na(max_length))) NA_real_ else max(max_length, na.rm = TRUE), 
                                        NA_real_),
            avg_density_m2        = mean(density_m2),
            avg_biomass_gm2       = mean(biomass_gm2),
            .groups = "drop")

biomass_by_station_and_taxa <- biomass_by_station_and_taxa |>
  filter(total_count > 0) |> 
  left_join(stations |> 
              select(ps_station_id, ps_site_id, exp_id, region, subregion, habitat, exposure, depth_strata, depth_m, divers, n_transects, survey_area_m2),
            by = "ps_station_id") |> 
  select(ps_station_id, ps_site_id, exp_id, 
         region, subregion, habitat, exposure, depth_strata, depth_m, divers, n_transects, survey_area_m2,
         accepted_name, accepted_aphia_id, rank, family, trophic_group,
         total_count, avg_count, avg_length_cm, min_length_cm, max_length_cm,
         avg_density_m2, avg_biomass_gm2)
```

```{r}
biomass_by_station_and_group <- biomass_by_station_and_taxa |> 
  group_by(ps_station_id, trophic_group) |> 
  summarize(avg_biomass_gm2 = round(sum(avg_biomass_gm2),2),
            .groups = "drop") |> 
  pivot_wider(names_from = trophic_group, values_from = avg_biomass_gm2, values_fill = 0) |> 
  pivot_longer(cols = -c(ps_station_id), names_to = "trophic_group", values_to = "avg_biomass_gm2") %>% 
  group_by(ps_station_id) %>% 
  mutate(pct_biomass = round(100 * avg_biomass_gm2 / sum(avg_biomass_gm2), 1)) |> 
  ungroup()
```

# 6. Summary Statistics

## Station Summary

Aggregate metrics at the station level 

```{r}
# Transect-level summaries

transect_summary <- biomass_by_transect_and_taxa |> 
  group_by(ps_station_id, diver, transect) |> 
  summarize(n_taxa      = n_distinct(accepted_aphia_id[count>0 & rank == "Species"]), 
            total_count = sum(count),
            density_m2  = sum(density_m2),
            biomass_gm2 = sum(biomass_gm2),
            .groups = "drop") 

# Average taxa richness per transect

avg_taxa_per_transect <- transect_summary |> 
  group_by(ps_station_id) |>
  summarise(avg_taxa = mean(n_taxa),
            .groups = "drop") 

# Station-level aggregation

station_stats <- biomass_by_station_and_taxa |> 
  group_by(ps_station_id) |> 
  summarize(n_taxa          = n_distinct(accepted_aphia_id[total_count > 0 & rank == "Species"]),
            total_count     = sum(total_count),
            avg_count       = sum(avg_count),
            avg_density_m2  = round(sum(avg_density_m2),3),
            avg_biomass_gm2 = round(sum(avg_biomass_gm2),3),
            .groups = "drop")

# % biomass by trophic group

biomass_pcts_by_station <- biomass_by_station_and_group %>%  
  select(-avg_biomass_gm2) %>% 
  pivot_wider(names_from = trophic_group, values_from = pct_biomass, names_prefix = "pct_", values_fill = 0) |> 
  janitor::clean_names() |> 
  rename(pct_lower_carni = pct_lower_carnivore)

# Final station summary

station_summary <- stations |> 
  left_join(station_stats, by = "ps_station_id") |> 
  left_join(avg_taxa_per_transect, by = "ps_station_id") |> 
  left_join(biomass_pcts_by_station, by = "ps_station_id")

station_summary <- station_summary |> 
  mutate(notes = NA_character_) 
```

```{r}
transect_summary |> 
  #filter(exposure != "lagoon") |> 
  #group_by(region) |> 
  summarise(n_stations         = n_distinct(ps_station_id),
            avg_n_taxa         = mean(n_taxa),
            mean_biomass_gm2   = mean(biomass_gm2),
            median_biomass_gm2 = median(biomass_gm2),
            across(contains("pct"), mean), 
            .groups = "drop") |> 
  mutate_if(is.numeric, round, 1) 
```


# 7. Map

Now, lets visualize the distribution of our sampling effort and summarize it by region and subregion to inspect any potential outliers.

```{r map}
#| label: fig-map
#| fig-cap: "Interactive map of BLT survey stations"

# Create spatial features

fish_sites_sf <- uvs_sites |> 
  select(ps_site_id, region, subregion, locality, habitat, exposure, latitude, longitude) |> 
  inner_join(
    stations |> 
      group_by(ps_site_id) |> 
      summarize(team = paste(unique(divers), collapse = "/"),
                strata = paste(unique(paste(depth_strata, " (", depth_m, "m)", sep = "")), 
                               collapse = "\n"),
                survey_area = sum(survey_area_m2),
                n_stations = n_distinct(ps_station_id),
                .groups = "drop"),
    by = "ps_site_id") |> 
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 

# Create interactive map
mapview::mapview(fish_sites_sf,
                 zcol = "exposure",
                 legend = TRUE,
                 map.types = "Esri.WorldImagery",
                 layer.name = "Exposure",
                 popup = leafpop::popupTable(fish_sites_sf, 
                                             zcol = c("ps_site_id", "strata", "team", "survey_area", "habitat", "exposure"))) |> 
  leafem::addMouseCoordinates() |> 
  addFullscreenControl()
```

# 8. Exploratory Visualizations

### missing data 

```{r}
#| label: fig-missingness-stations
#| fig-cap: "Missingness plot for station summary"
#| 
naniar::vis_miss(station_summary)
```

### by site

```{r}
biomass_by_station_and_group |> 
  left_join(stations %>% 
              select(ps_site_id, depth_strata, region, ps_station_id), by = "ps_station_id") %>% 
  group_by(region, ps_site_id, depth_strata, trophic_group) |> 
  summarize(avg_biomass_gm2 = round(sum(avg_biomass_gm2),2),
            .groups = "drop") |> 
  mutate(site = str_extract(ps_site_id, "[^_]+$")) |> 
  filter(depth_strata != "supershallow") |> 
  ggplot()+
  geom_col(aes(site, avg_biomass_gm2, fill = trophic_group))+
  facet_grid(depth_strata ~ region, scales = "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  labs(x = NULL, y = "Fish biomass (g/m²)",
       fill = "",
       title = paste(exp_id, ":","Fish biomass composition by site, region, and depth"))

ggsave(filename = file.path(figures_out, paste0(exp_id, "_blt_biomass_by_site.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

### by subregion

```{r}
biomass_by_station_and_group |> 
  left_join(stations %>% select(ps_site_id, depth_strata, region, subregion, ps_station_id), by = "ps_station_id") %>% 
  group_by(region, subregion, depth_strata, trophic_group) |> 
  summarize(avg_biomass_gm2 = mean(avg_biomass_gm2)) |> 
  ungroup() |> 
  filter(depth_strata != "supershallow") |> 
  ggplot()+
  geom_col(aes(subregion, avg_biomass_gm2, fill = trophic_group))+
  facet_grid(depth_strata~region, scales = "free")+
  theme_pristineseas()+
  labs(x = NULL, y = "Fish biomass (g/m²)",
       fill = "",
       title = paste(exp_id,":","Fish biomass composition by subregion, and depth"))+
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"),
        legend.position = "bottom",
        legend.key.height = unit(4, "mm"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(filename = file.path(figures_out, paste0(exp_id, "_blt_biomass_by_subregion.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

## by strata

Examine how key ecological metrics vary across depth strata to understand vertical zonation patterns.

```{r}
#| label: fig-transects-summary
#| fig-cap: "Comparison of species richness, total count, density, and biomass across depth strata"

station_summary |> 
  select(ps_station_id, depth_strata, avg_taxa, avg_count, avg_density_m2, avg_biomass_gm2) |>
  pivot_longer(cols = c("avg_taxa", "avg_count", "avg_density_m2", "avg_biomass_gm2"),
               names_to = "metric",
               values_to = "value") |> 
  mutate(metric = recode(metric, 
                         "avg_taxa"        = "Species Richness",
                         "avg_count"       = "Count",
                         "avg_density_m2"  = "Density (ind/m²)",
                         "avg_biomass_gm2" = "Biomass (g/m²)")) |> 
  ggplot(aes(x = depth_strata, y = value, fill = depth_strata)) +
  geom_boxplot(alpha = 0.8, color = "gray40", outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.5, color = "gray30") +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "", 
       y = "",
       fill = "",
       title = "Transect Metrics by Depth Strata",
       subtitle = paste(exp_id, ":","Comparison of species richness, total count, density, and biomass across depth strata")) +
  scale_fill_manual(values = ps_colors("depth_strata"))

ggsave(filename = file.path(figures_out, paste0(exp_id, "_blt_station_metrics_by_depth.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

### prevalent taxa 

```{r}
#| label: fig-top-taxa
#| fig-cap: "Biomass distributions for top taxa and depth strata"
#| warning: false
#| fig-width: 12
#| fig-height: 8

# Fill missing combinations so absences are treated as zeros
plot_df <- biomass_by_station_and_taxa |> 
  select(ps_station_id, accepted_name, avg_biomass_gm2) |> 
  complete(ps_station_id, 
           accepted_name, 
           fill = list(avg_biomass_gm2 = 0)) |> 
  mutate(accepted_name = forcats::fct_reorder(accepted_name, avg_biomass_gm2, .fun = median, .desc = TRUE)) 

# Identify major taxa by total biomass
taxa_means <- plot_df |> 
  group_by(accepted_name) |> 
  summarise(mean_biomass   = mean(avg_biomass_gm2, na.rm = TRUE),
            median_biomass = median(avg_biomass_gm2, na.rm = TRUE)) |> 
  arrange(desc(median_biomass), desc(mean_biomass)) 

top_taxa <- taxa_means |> 
  slice_head(n = 12) 

# Summarize by depth for medians and means
depth_summary <- plot_df |> 
  left_join(stations |> distinct(ps_station_id, depth_strata), by = "ps_station_id") |> 
  filter(accepted_name %in% top_taxa$accepted_name) |> 
  group_by(accepted_name, depth_strata) |> 
  summarise(mean_biomass_gm2   = mean(avg_biomass_gm2, na.rm = TRUE),
            median_biomass_gm2 = median(avg_biomass_gm2, na.rm = TRUE),
            n_obs = n(),
            .groups = "drop") 

# Plot distributions by depth and taxon
plot_df |> 
  left_join(stations |> select(ps_station_id, depth_strata), by = "ps_station_id") |> 
  filter(accepted_name %in% top_taxa$accepted_name) |> 
  ggplot(aes(x = depth_strata, y = avg_biomass_gm2)) +
  geom_boxplot(aes(color = accepted_name),
               outlier.shape = NA, fill = "white",
               alpha = 0.3, linewidth = 0.6, show.legend = FALSE) +
  geom_jitter(aes(color = accepted_name),
              width = 0.25, size = 1.5, alpha = 0.5, show.legend = FALSE) +
  geom_point(data = depth_summary,
             aes(y = mean_biomass_gm2),
             shape = 23, fill = "black", color = "white",
             size = 3, stroke = 0.8) +
  facet_wrap(~ accepted_name, scales = "free_y", ncol = 4) +
  scale_y_log10(breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
               labels = c("0.001", "0.01", "0.1", "1", "10", "100"),
               expand = expansion(mult = c(0.05, 0.15))) +
  paletteer::scale_color_paletteer_d("ggsci::default_igv") +
  theme_pristineseas(base_family = "Merriweather") +
  theme(strip.text = element_text(size = 10, face = "italic", lineheight = 1.2, margin = margin(b = 8))) +
  labs(x = NULL,
       y = "Biomass (g/m², log scale)",
       title = "Do dominant taxa show distinct depth preferences?",
       subtitle = "Top 12 taxa by median biomass • Diamonds = mean • Taxa ordered by overall biomass",
       caption = "Y-axis on log scale; zeros filled via completion step")
```

### biomass vs. density

Examine the relationship between biomass and density across trophic groups to understand energy flow.

```{r}
#| label: fig-points-by-taxa
#| fig-cap: "Biomass vs. Density Relationship by Trophic Group"
#| fig-width: 7
#| fig-height: 6

biomass_by_station_and_taxa |> 
  filter(total_count > 0) |> 
  ggplot(aes(x = avg_density_m2, y = avg_biomass_gm2, color = trophic_group)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_density_2d(color = "black", linewidth = 0.2) + # Density contours for highlighting structure
  scale_y_log10(labels = scales::comma_format(accuracy = 0.01)) +
  scale_x_sqrt(labels = scales::comma_format(accuracy = 0.01)) +
  labs(x = "Density (ind/m², sqrt scale)", 
       y = "Biomass (g/m², log scale)", 
       color = "",
       title = "Biomass vs. Density Relationship",
       subtitle = "Log-transformed biomass vs. square-root-transformed density across trophic groups") +
  theme_pristineseas(base_family = "Merriweather") 
```

---

# 9. Export & Upload

## Pre-upload summary

```{r}
#| label: data-validation
#| echo: true

# Validate data completeness and basic consistency
validation_summary <- list(
  stations = list(
    total_records      = nrow(station_summary),
    complete_records   = sum(complete.cases(station_summary |> select(-locality, -notes))),
    missing_critical   = sum(is.na(station_summary$ps_station_id) |
                             is.na(station_summary$exp_id) |
                             is.na(station_summary$region)),
    n_regions          = n_distinct(station_summary$region),
    n_subregions       = n_distinct(station_summary$subregion),
    n_depth_strata     = n_distinct(station_summary$depth_strata)
  ),
  
  observations = list(
    total_records       = nrow(biomass_per_obs),
    complete_records    = sum(complete.cases(biomass_per_obs)),
    species_count       = n_distinct(biomass_per_obs$accepted_name),
    largest_fish_cm     = max(biomass_per_obs$length_cm, na.rm = TRUE),
    largest_fish_taxa   = biomass_per_obs$accepted_name[which.max(biomass_per_obs$length_cm)],
    smallest_fish_cm    = min(biomass_per_obs$length_cm, na.rm = TRUE),
    smallest_fish_taxa  = biomass_per_obs$accepted_name[which.min(biomass_per_obs$length_cm)],
    negative_lengths    = sum(biomass_per_obs$length_cm < 0, na.rm = TRUE)
  ),
  
  biomass_summaries = list(
    max_biomass_gm2        = max(station_summary$avg_biomass_gm2, na.rm = TRUE),
    max_biomass_station    = station_summary$ps_station_id[which.max(station_summary$avg_biomass_gm2)],
    max_biomass_subregion  = station_summary$subregion[which.max(station_summary$avg_biomass_gm2)],
    min_biomass_gm2        = min(station_summary$avg_biomass_gm2, na.rm = TRUE),
    min_biomass_station    = station_summary$ps_station_id[which.min(station_summary$avg_biomass_gm2)],
    min_biomass_subregion  = station_summary$subregion[which.min(station_summary$avg_biomass_gm2)],
    zero_biomass_stations  = sum(station_summary$avg_biomass_gm2 == 0, na.rm = TRUE)
  )
)

# Display validation summary
cat("\n📊 DATA VALIDATION SUMMARY\n")
cat("==========================\n\n")

cat("Station Summary:\n")
cat("  • Total records:          ", validation_summary$stations$total_records, "\n")
cat("  • Complete records:       ", validation_summary$stations$complete_records, "\n")
cat("  • Missing critical fields:", validation_summary$stations$missing_critical, "\n")
cat("  • Regions:                ", validation_summary$stations$n_regions, 
    " | Subregions:", validation_summary$stations$n_subregions,
    " | Depth strata:", validation_summary$stations$n_depth_strata, "\n\n")

cat("Observations:\n")
cat("  • Total records:          ", validation_summary$observations$total_records, "\n")
cat("  • Complete records:       ", validation_summary$observations$complete_records, "\n")
cat("  • Unique species:         ", validation_summary$observations$species_count, "\n")
cat("  • Largest fish:           ", validation_summary$observations$largest_fish_cm, " cm (",
    validation_summary$observations$largest_fish_taxa, ")\n")
cat("  • Smallest fish:          ", validation_summary$observations$smallest_fish_cm, " cm (",
    validation_summary$observations$smallest_fish_taxa, ")\n")

cat("Biomass Summaries:\n")
cat("  • Max biomass:            ", validation_summary$biomass_summaries$max_biomass_gm2, " g/m² (",
    validation_summary$biomass_summaries$max_biomass_station, ", ",
    validation_summary$biomass_summaries$max_biomass_subregion, ")\n")
cat("  • Min biomass:            ", validation_summary$biomass_summaries$min_biomass_gm2, " g/m² (",
    validation_summary$biomass_summaries$min_biomass_station, ", ",
    validation_summary$biomass_summaries$min_biomass_subregion, ")\n")
cat("  • Stations with zero biomass:", validation_summary$biomass_summaries$zero_biomass_stations, "\n\n")

cat("✅ Basic validation complete.\n")
```

## Schema Validation

```{r}
validate_and_export <- function(df, table_ref, label, filename) {
  
  ref_cols <- tbl(bq_connection, table_ref) |> 
    filter(!is.na(exp_id)) |> 
    head() |> 
    collect() |> 
    names()
  
  missing <- setdiff(ref_cols, names(df))
  extra <- setdiff(names(df), ref_cols)
  
  if (length(missing) > 0 || length(extra) > 0) {
    cat("\n❌ Schema mismatch for", label, "\n")
    if (length(missing)) cat("  Missing columns:", paste(missing, collapse = ", "), "\n")
    if (length(extra)) cat("  Extra columns:", paste(extra, collapse = ", "), "\n")
    stop("Fix schema before export", call. = FALSE)
  }
  
  out_path <- file.path(data_out, filename)
  df |> select(all_of(ref_cols)) |> write_csv(out_path)
  cat("✓", label, "→", out_path, "\n")
  invisible(out_path)
}
```

## CSV Exports

```{r}
stations_csv <- validate_and_export(df        = station_summary |> 
                                      mutate(pct_top_predator = NA, pct_shark = NA, pct_herbi_detri = NA, pct_plankti = NA),
                                    table_ref = "uvs.blt_stations",
                                    label     = "Stations",
                                    filename  = paste0(exp_id, "_uvs_blt_stations.csv"))

observations_csv <- validate_and_export(df        = biomass_per_obs |> 
                                          mutate(notes = NA_character_,
                                                 terminal_phase = FALSE,
                                                 station_label = stratify(depth_m)) ,
                                        table_ref = "uvs.blt_observations",
                                        label     = "Observations",
                                        filename  = paste0(exp_id, "_uvs_blt_observations.csv"))

biomass_csv <- validate_and_export(df        = biomass_by_station_and_taxa,
                                   table_ref = "uvs.blt_biomass_by_station_taxa",
                                   label     = "Biomass by Station | Taxa",
                                   filename  = paste0(exp_id, "_uvs_blt_biomass_by_station_taxa.csv"))
```