---
title: "Expedition Sites"
date: today
format: 
  html:
    theme: minty
    self-contained: true
    code-fold: true
    toc: true 
    toc-depth: 3
    toc-location: right
execute:
  fig-width: 10
---

```{r setup, message = F, warning = F, fig.width = 10, fig.height = 10, echo = F}
options(scipen = 999)

# Load libraries

library(tidyverse)
library(readxl)
library(janitor)
library(PristineSeasR2)
library(hms)
library(pointblank)
library(sf)
library(mapview)
library(gt)
library(bigrquery)
library(basemaps)
library(ggrepel)
library(showtext)
library(ggspatial)

# Add Google font for consistent visual outputs

font_add_google("Lato", "lato")
showtext_auto()

# =============================================================================
# Expedition Configuration
# =============================================================================

exp_config <- list(exp_id      = "NIU_2016",
                   date_bounds = c(as.Date("2016-01-01"), as.Date("2016-12-31")),
                   lat_min     = -12.3, 
                   lat_max     = -5,  
                   lon_min     = 155, 
                   lon_max     = 170)  


# Setup paths and BQ auth

ps_paths <- PristineSeasR2::get_drive_paths()

exp_path <- file.path(ps_paths$expeditions, str_replace(exp_config$exp_id, "_", "-"))

bigrquery::bq_auth(email = "marine.data.science@ngs.org")

bq_connection <- DBI::dbConnect(bigrquery::bigquery(), project = "pristine-seas")

exp_id <- exp_config$exp_id

legacy_path <- "/Users/marinedatascience/Library/CloudStorage/GoogleDrive-marine.data.science@ngs.org/.shortcut-targets-by-id/16vjFvzPNgx8MuRqNqZj-Uqc08bkAVKqX/Pristine Seas/SCIENCE/legacy data/Niue/"
```

## Overview

This script processes and validates **Underwater Visual Survey (UVS) site metadata** for the expedition **`r exp_id`**. 

**Workflow**: Import → Standardize → QA/QC → Visualize → Export → Upload

This process ensures data integrity prior to analysis and database ingestion.

## 1. Import and Standardize Fieldbook Data

```{r uvs-import}
# 1. Import Underwater Visual Surveys (UVS) fieldbook

uvs_sites <- read_xlsx(file.path(legacy_path, "Fish", "Niue_fish_data_AMF.xlsx"),
                       sheet = "metadata") |> 
  clean_names() |> 
  rename(latitude  = lat,
         longitude = long,
         region    = reef,
         site      = station) |> 
  mutate(exp_id    = exp_id,
         method    = "uvs",
         date      = lubridate::ymd(date),
         time      = hms::as_hms(time),
         habitat   = str_to_lower(habitat),
         exposure  = str_to_lower(exposure),
         region    = str_to_title(region),
         subregion = region,
         locality  = NA_character_,
         in_mpa    = if_else(mangt == "MPA", TRUE, FALSE),
         notes     = NA_character_,
         site_name = NA_character_,
         team_lead = "Alan Friedlander",
         # Standardize site IDs to match database format (3-digit padding)
         ps_site_id =  paste0("NIU_2016_uvs_", str_pad(site, 3, pad = "0"))) |> 
  # Arrange columns in logical order
  select(ps_site_id, exp_id, method, region, subregion, locality, 
         date, time, latitude, longitude, habitat, exposure, site_name, in_mpa, team_lead, notes)

uvs_sites$habitat[uvs_sites$habitat == "forereef"] <- "fore_reef"
uvs_sites$habitat[uvs_sites$habitat == "backreef"] <- "back_reef"

# Create regions lookup from actual data for validation

regions_lookup_table <- uvs_sites |> 
  distinct(region, subregion) 

valid_regions <- unique(regions_lookup_table$region)
valid_subregions <- unique(regions_lookup_table$subregion)

# Need to load exposures separately

exps <- read_xlsx(file.path(legacy_path, "Fish", "Exposure.xlsx")) |> 
  clean_names() |> 
  transmute(ps_site_id =  paste0("NIU_2016_uvs_", str_pad(station, 3, pad = "0")),
            exposure   = case_when(exposure == "Wind" ~ "windward",
                                   exposure == "Lee" ~ "leeward",
                                   TRUE ~ NA_character_))

uvs_sites <- uvs_sites |> 
  select(-exposure) |> 
  left_join(exps)

uvs_sites$exposure[uvs_sites$habitat == "back_reef"] <- "lagoon"
```

We conducted a total of `r nrow(uvs_sites)` UVS surveys across `r length(unique(uvs_sites$region))` regions and `r length(unique(uvs_sites$subregion))` subregions. The data includes `r sum(uvs_sites$in_mpa, na.rm = TRUE)` sites within Marine Protected Areas (MPAs). The first site was surveyed on `r min(uvs_sites$date)` and the last on `r max(uvs_sites$date)`.

## 2. QA/QC and Validation

```{r uvs_qaqc}
# 2.QA/QC using pointblank

uvs_sites_qaqc <- uvs_sites |> 
  create_agent(label = "UVS sites QA/QC", tbl_name = "uvs_sites_raw") |> 
  # Ensure each site ID is unique
  rows_distinct(ps_site_id,
                label = "Unique site IDs",
                actions = action_levels(stop_at = 0.001)) |> 
  # Enforce non-missing values for critical columns
  rows_complete(columns = vars(ps_site_id, latitude, longitude, date),
                label = "Complete rows for key fields",
                actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |>
  # Habitat and exposure validation against allowed vocab
  col_vals_in_set(columns = vars(habitat),
                  set = allowed_vocab$uvs_habitats,
                  label = "Valid habitat values",
                  actions = action_levels(stop_at = 0.1)) |>
  col_vals_in_set(columns = vars(exposure),
                  set = allowed_vocab$exposure,
                  label = "Valid exposure values",
                  actions = action_levels(stop_at = 0.1)) |>
  # Region validation against lookup table
  col_vals_in_set(columns = vars(region),
                  set = valid_regions,
                  label = "Region in lookup table",
                  actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |>
  col_vals_in_set(columns = vars(subregion), 
                  set = valid_subregions,
                  label = "Subregion in lookup table",
                  actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |>
  # Geographic bounds validation
  col_vals_between(columns = vars(latitude),
                   left = exp_config$lat_min,
                   right = exp_config$lat_max,
                   label = "Latitude within bounds",
                   actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |>
  col_vals_between(columns = vars(longitude),
                   left = exp_config$lon_min, 
                   right = exp_config$lon_max,
                   label = "Longitude within bounds",
                   actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |> 
  # Date within expedition range
  col_vals_between(columns = vars(date),
                   left = exp_config$date_bounds[1],
                   right = exp_config$date_bounds[2], 
                   label = "Date within expedition period",
                   actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |> 
  # Critical NAs
  col_vals_not_null(columns = vars(latitude, longitude),
                    label = "Coordinates are not missing",
                    actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |>
  col_vals_not_null(columns = vars(date),
                    label = "Date is not missing", 
                    actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |> 
  interrogate()


# Display QA/QC results
uvs_sites_qaqc
```


Validation checks confirmed that all site IDs are unique and coordinates fall within the expedition bounds. 

```{r uvs-missing-data}
#| fig-cap: "Missing data in UVS sites fieldbook"

naniar::vis_miss(uvs_sites)
```

## 3. Visualize

The interactive map below displays unique UVS sites with metadata-rich popups.

```{r uvs-map}
# Interactive visual map

uvs_sites_sf <- uvs_sites |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  distinct(ps_site_id, region, subregion, locality, habitat, exposure, in_mpa, geometry)

mapview::mapview(uvs_sites_sf,
                 zcol = "exposure",
                 legend = TRUE,
                 map.types = "Esri.WorldImagery",
                 layer.name = "in MPA",
                 popup = leafpop::popupTable(uvs_sites_sf, 
                                             zcol = c("ps_site_id", "region", "subregion", "locality", "habitat", "exposure", "in_mpa"))) |> 
  leafem::addMouseCoordinates() |> 
  leaflet.extras::addFullscreenControl()
```

```{r eval = F, include = F}
plot_uvs_sites_map <- function(region, subregion = NULL, save = FALSE, path = NULL) {

  # Filter and transform
  map_data <- uvs_sites_sf |>
  mutate(site_number = str_extract(ps_site_id, "\\d+(?!.*\\d)")) |>
  filter(region == !!region)

if (!is.null(subregion)) {
  map_data <- filter(map_data, subregion == !!subregion)
}
map_data <- st_transform(map_data, 3857)

  # Coordinates for label repel
  coords <- st_coordinates(map_data)
  
  map_data_coords <- map_data |>
    mutate(x = coords[, 1], y = coords[, 2])

  # Bounding box + buffer
  bbox <- st_bbox(map_data)
  
  buffer_meters <- 1000
  
  bbox_buffered_m <- c(xmin = bbox[1] - buffer_meters,
                       ymin = bbox[2] - buffer_meters,
                       xmax = bbox[3] + buffer_meters,
                       ymax = bbox[4] + buffer_meters)
  
  class(bbox_buffered_m) <- "bbox"
  
  attr(bbox_buffered_m, "crs") <- st_crs(3857)

  # Basemap
  aoi <- basemap_ggplot(bbox_buffered_m,
                        map_service = "esri",
                        map_type = "world_imagery",
                        map_res = 1)

  # Compose title/subtitle
  subtitle_text <- if (!is.null(subregion)) { 
    paste(region, "|", subregion)} else {
      region}

  # Final plot
  p <- aoi +
    geom_sf(data = map_data, shape = 21, size = 4,
            fill = "yellow", color = "black", stroke = 0.8, alpha = 0.85) +
    annotation_scale(location = "bl", width_hint = 0.3,
                     pad_x = unit(0.5, "cm"), pad_y = unit(0.5, "cm")) +
    annotation_north_arrow(location = "tr", which_north = "true",
                           pad_x = unit(0.2, "cm"), pad_y = unit(0.2, "cm"),
                           style = north_arrow_fancy_orienteering) +
    labs(title = "Underwater Survey Sites",
         subtitle = subtitle_text,
         x = NULL, y = NULL) +
    scale_x_continuous(breaks = seq(-180, 180, by = 0.025), expand = c(0, 0)) +
    scale_y_continuous(breaks = seq(-90, 90, by = 0.025), expand = c(0, 0)) +
    theme_minimal(base_family = "lato") +
    theme(plot.title = element_text(size = 16, face = "bold", hjust = 0),
          plot.subtitle = element_text(size = 14, hjust = 0, color = "gray30"),
          axis.text = element_text(size = 12, color = "gray30"),
          panel.grid.major = element_line(color = "gray85", linewidth = 0.2),
          plot.margin = margin(10, 10, 10, 10))
  
  # Save if needed
  if (isTRUE(save)) {
    if (is.null(path)) {
      stop("Please provide a file path to save the plot.")
    }
    ggsave(path, plot = p, width = 10, height = 10, dpi = 300)
  }
  
  return(p)
}

plot_uvs_sites_map(region = "Niue", subregion = NULL,
                   save = TRUE, path = file.path(exp_path, "figures/site_maps/", "uvs_Niue.pdf"))
```

## 4. Summary and Export

Of the `r nrow(uvs_sites)` sites, `r sum(uvs_sites$in_mpa, na.rm = TRUE)` are located within MPAs. Surveys covered `r length(unique(uvs_sites$habitat))` habitat types.

```{r}
#| label: tbl-uvs-sites-by-habitat
#| tbl-cap: "Summary of UVS survey effort by region, exposure, and habitat."

uvs_sites |> 
  group_by(region, subregion, exposure) |>
  summarize(n_surveys = n(),
            .groups = "drop") |> 
  pivot_wider(names_from = exposure, values_from = n_surveys, values_fill = 0) |> 
  gt::gt(groupname_col = "leg") |> 
  gt::tab_options(row_group.as_column = T,
                  table.width = pct(80),
                  table.font.size = 12,
                  table.border.top.style = "solid",
                  table.border.top.width = px(1),
                  table.border.top.color = "black",
                  heading.align = "center",
                  column_labels.font.weight = "bold")
```

```{r eval = F}
bq_table_upload(bq_table("pristine-seas", "uvs", "sites"),
                values = uvs_sites,
                create_disposition = "CREATE_NEVER",
                write_disposition = "WRITE_APPEND")
```

