---
title: "Benthos LPI - Data Pipeline"
date: today
format: 
  html:
    theme: minty
    self-contained: true
    code-fold: true
    toc: true 
    toc-depth: 3
    toc-location: right
    html-table-processing: none
execute:
  fig-width: 10
---

```{r setup, message = F, warning = F, fig.width = 10, fig.height = 10, echo = F}
options(scipen = 999)

# Hook to format inline numeric expressions with comma separators:
knitr::knit_hooks$set(inline = function(x) {
  # For non-numeric values, just return as character
  if (!is.numeric(x)) {return(as.character(x))}
  # Format numbers with comma as big.mark, no scientific notation
  else format(x, big.mark = ",", scientific = FALSE)})

library(sf)
library(hms)
library(readxl)
library(janitor)
library(lubridate)
library(gt)
library(pointblank)
library(tidyverse)
library(bigrquery)
library(leaflet)
library(leaflet.extras)
library(ggtext)
library(PristineSeasR2)

# ---- Parameters -----------------------------------------------------------
exp_id   <- "NIU_2016"

ps_paths <- PristineSeasR2::get_drive_paths()

prj_path <- file.path(ps_paths$projects, "legacy-db")

# ---- Auth & Connection ----------------------------------------------------
bigrquery::bq_auth(email = "marine.data.science@ngs.org")
bq_connection <- DBI::dbConnect(bigrquery::bigquery(), project = "pristine-seas")

set_theme(PristineSeasR2::theme_pristineseas())

data_out <- file.path(prj_path, "data/output/uvs")

figures_out <- file.path(prj_path, "figures/uvs")

legacy_path <- file.path(ps_paths$science, "legacy data/Niue/")
```

# Overview

This script outlines the end-to-end pipeline for processing Line Point Intercept (LPI) survey data collected during Pristine Seas expeditions. The pipeline ingests raw field data, standardizes formats, performs taxonomy lookups, applies quality assurance/quality control (QA/QC), computes station-level summaries, explores pattenrs, and loads the clean data into the Pristine Seas Science Database in BigQuery.

# 1. Data Ingestion

## Reference Tables

We begin by reading the UVS (Underwater Visual Survey) site information already processed.
  
```{r load_references}
uvs_sites <- tbl(bq_connection, "uvs.sites") |> 
   filter(exp_id == !!exp_id) |> 
   collect()

# Validate site data
cat("Sites loaded:", nrow(uvs_sites), "\n")
cat("Regions:",      unique(uvs_sites$region), "\n")
cat("Subregions:",   unique(uvs_sites$subregion), "\n")

# Benthic taxonomy reference
benthos_taxa_lut <- tbl(bq_connection, "taxa_info.benthos") |> 
  filter(!is.na(accepted_name)) |> 
  collect()

cat("✓ Loaded", nrow(benthos_taxa_lut), "benthic taxa\n")
```

## Fieldbooks

Next, process field observations from different divers, ensuring consistent formatting and traceability.

  - Field names are harmonized, and units are standardized.
  - Unique observation IDs (obs_id) are generated for QA/QC.
  - Field data is cleaned and formatted to ensure database compatibility.

```{r}
sections_raw <- read_xlsx(file.path(legacy_path, "Benthic/Science Kike/Transsects-Niue.xlsx"),
                          sheet         = "Rough data",
                          range         = "D1:MU2",
                          col_names     = FALSE,
                          .name_repair  = "minimal") |>
  t() |>
  as_tibble(.name_repair = "minimal") |> 
  row_to_names(row_number = 1) |>
  clean_names()

sections <- sections_raw |> 
  mutate(depth_m          = parse_number(depth),
         site             = str_pad(parse_number(sample), 3, pad = "0"),
         transect         = str_remove(sample, "\\d+"),
         ps_site_id       = paste(exp_id, "uvs", site, sep = "_"),
         transect_section = rep(c("0-10", "10-20", "20-30", "30-40", "40-50"), 
                                n_distinct(sample))) |> 
  select(ps_site_id, depth_m, transect, transect_section) |> 
  group_by(ps_site_id, transect) |>
  mutate(station_depth_m = mean(depth_m, na.rm = TRUE),
         n_sections = n()) |> 
  ungroup() |> 
  mutate(exp_id         = exp_id,
         survey_type    = "uvs",
         method         = "lpi",
         diver          = "Kike Ballesteros",
         depth_strata   = stratify(station_depth_m),
         ps_station_id  = paste(ps_site_id, station_suffix(depth_strata), sep = "_"),
         section_id     = paste(ps_station_id, transect, transect_section, sep = "_")) |> 
  select(exp_id, survey_type, ps_site_id, ps_station_id, method, diver, 
         depth_m, station_depth_m, depth_strata, transect, 
         transect_section, section_id)

contacts_raw <- read_xlsx(file.path(legacy_path, "Benthic/Science Kike/Transsects-Niue.xlsx"),
                          sheet         = "Rough data",
                          range         = "C6:MU59",
                          col_names     = FALSE,
                          .name_repair  = "minimal") 

valid_ids <- sections$section_id

contacts <- contacts_raw |>
    set_names(c("morphotaxon","field_name", valid_ids)) |>
    select(morphotaxon, field_name, all_of(valid_ids)) |>
    mutate(across(all_of(valid_ids), as.character)) |>
    pivot_longer(cols      = -c(field_name, morphotaxon),  # Exclude BOTH taxonomy columns
                 names_to  = "section_id",
                 values_to = "contacts") |>
    mutate(contacts = parse_number(contacts)) |>
    filter(!is.na(contacts)) |>
    select(section_id, field_name, morphotaxon, contacts)

section_totals <- contacts |>
    group_by(section_id) |>
    summarise(n_points = sum(contacts, na.rm = TRUE),
              .groups  = "drop")
  
lpi_sections <- sections |>
  left_join(section_totals, by = "section_id")

lpi_contacts <- contacts |>
  left_join(select(sections, section_id, exp_id, ps_station_id, transect, transect_section, diver, depth_m), 
            by = "section_id") |>
  select(ps_station_id, exp_id, diver, transect, depth_m, transect_section, morphotaxon, field_name, contacts, section_id) |>
  arrange(ps_station_id, transect, transect_section) 
```

# 2. Data Transformation

```{r}
# Explore section aligment

lpi_contacts <- lpi_contacts |> 
  mutate(field_name = if_else(field_name == "Tubastrea", "Hard coral", field_name)) |> 
  group_by(ps_station_id, exp_id, diver, depth_m, transect, transect_section, morphotaxon, field_name, section_id) |>
  summarize(count = sum(contacts), .groups = "drop")

lpi_sections <- lpi_sections |> 
  left_join(lpi_contacts |>
              group_by(section_id) |>
              summarise(lpi_coral_pts = sum(count[morphotaxon == "Hard coral"], na.rm = TRUE),
                        .groups       = "drop")) |> 
  mutate(lpi_pct_coral = round(100 * lpi_coral_pts / n_points, 2))

total_lpi_pts <- sum(lpi_contacts$count)
total_lpi_coral_pts <- sum(lpi_contacts$count[lpi_contacts$morphotaxon == "Hard coral"])

total_lpi_coral_pts/total_lpi_pts
```

## Merging LPI and Coral data

Now, we merge the broad LPI data with the hard coral contacts from the coral specialist. In theory, divers follow the same transect line and identify taxa at the same contact points, but inevitably, factors such as currents, surge, and diver bias can lead to differences in the number of hard coral contacts recorded. The goal is to reconcile these differences while preserving the integrity of the systematic LPI sampling framework.

We accomplish this by:

  - Trusting the LPI totals - The systematic LPI method provides unbiased estimates of overall hard coral abundance
  - Leveraging coral specialist expertise - The coral diver provides high-resolution taxonomic identification
  - Proportional apportionment - We distribute the LPI hard coral total among coral species based on the proportions observed by the coral specialist

Edge case handling:

  - Normal case (both divers recorded hard coral): LPI hard coral total is apportioned among coral species based on specialist's observed proportions
  - Incomplete transect (LPI recorded hard coral, but coral specialist has no data): Hard coral remains as generic "Scleractinian" category, indicating the coral specialist was unable to complete identification for that section
  - Methodological discrepancy / Off-transect observations (coral specialist found species but LPI recorded zero hard coral): Species records are classified as off-transect observations with transect_section = "Off" and aggregated at the station level. This preserves species occurrence data while maintaining the integrity of systematic transect sampling by clearly separating opportunistic observations from quantitative abundance estimates.

This approach maintains the statistical rigor of the LPI methodology while gaining the taxonomic resolution needed for coral community analysis. The resulting dataset preserves both the overall hard coral coverage (from LPI) and the species composition (from coral specialist), providing the best of both sampling approaches.

```{r}
coral_sections <- readxl::read_xlsx(file.path(legacy_path, "Benthic/Benthos_Corals_Niue.xlsx"),
                  col_names = FALSE,
                  sheet = "Niue_Abundance",
                  range = "B3:GS6") %>%
  janitor::clean_names() |> 
  t() %>% 
  as_tibble() |> 
  set_names(c("site", "access","exposure", "depth_m")) |> 
  mutate(depth_m          = parse_number(depth_m),
         site             = str_pad(parse_number(site), 3, pad = "0"),
         ps_site_id       = paste(exp_id, "uvs", site, sep = "_")) |> 
  select(ps_site_id, depth_m)

coral_sections_2 <- readxl::read_xlsx(file.path(legacy_path, "Benthic/Benthos_Corals_Niue.xlsx"),
                                    sheet = "Beveridge_Abundance",
                  col_names = FALSE,
                  range = "B3:EZ4") %>%
  janitor::clean_names() |> 
  t() %>% 
  as_tibble() |> 
  set_names(c("site",  "depth_m")) |> 
  mutate(depth_m          = as.numeric(depth_m),
         site             = str_pad(site, 3, pad = "0"),
         ps_site_id       = paste(exp_id, "uvs", site, sep = "_")) |> 
  select(ps_site_id, depth_m)

coral_sections <- bind_rows(coral_sections, coral_sections_2)

sections_names <- c("0-10", "10-20", "20-30", "30-40", "40-50")

coral_sections <- coral_sections |> 
  group_by(ps_site_id) |> 
  mutate(transect_section = rep(rev(sections_names), n_distinct(depth_m))) |> 
  ungroup() |> 
  select(ps_site_id, depth_m, transect_section)

coral_sections <- coral_sections |> 
  left_join(coral_sections |> 
              distinct(ps_site_id, depth_m) |> 
              group_by(ps_site_id) |> 
              mutate(transect = LETTERS[row_number()])) |> 
  group_by(ps_site_id, transect) |>
  mutate(station_depth_m = mean(depth_m, na.rm = TRUE),
         n_sections = n()) |> 
  ungroup() |> 
  # Wrangle other fields
  mutate(exp_id         = exp_id,
         survey_type    = "uvs",
         method         = "lpi",
         diver          = "Eric Brown",
         depth_strata   = stratify(station_depth_m),
         ps_station_id  = paste(ps_site_id, station_suffix(depth_strata), sep = "_"),
         section_id     = paste(ps_station_id, transect, transect_section, sep = "_")) |> 
  select(exp_id, survey_type, ps_site_id, ps_station_id, method, diver, 
         depth_m, station_depth_m, depth_strata, transect, 
         transect_section, section_id)

tmp <- coral_sections |> 
  left_join(uvs_sites |> 
              select(ps_site_id, region))

sec_id_niue <- tmp$section_id[tmp$region == "Niue"]
sec_id_bev <- tmp$section_id[tmp$region == "Beveridge"]

# Now read the coral contacts 
## Niue

coral_contacts_niue <- readxl::read_xlsx(file.path(legacy_path, "Benthic/Benthos_Corals_Niue.xlsx"),
                                         col_names = FALSE,
                                         sheet = "Niue_Abundance",
                                         range = "A7:GS97") %>%
  janitor::clean_names() |> 
  set_names(c("morphotaxon", sec_id_niue))|> 
  mutate(across(all_of(sec_id_niue), as.character)) |> 
  pivot_longer(cols      = -c(morphotaxon),  # Exclude BOTH taxonomy columns
                 names_to  = "section_id",
                 values_to = "contacts") |>
    mutate(count = parse_number(contacts)) |>
    filter(!is.na(count)) |>
    select(section_id, morphotaxon, count)

## Beve

coral_contacts_bev <- readxl::read_xlsx(file.path(legacy_path, "Benthic/Benthos_Corals_Niue.xlsx"),
                                        col_names = FALSE,
                                        sheet = "Beveridge_Abundance",
                                        range = "A5:EZ77") %>%
  janitor::clean_names() |> 
  set_names(c("morphotaxon", sec_id_bev)) |> 
  mutate(across(all_of(sec_id_bev), as.character)) |> 
  pivot_longer(cols      = -c(morphotaxon),  # Exclude BOTH taxonomy columns
                 names_to  = "section_id",
                 values_to = "contacts") |>
    mutate(count = parse_number(contacts)) |>
    filter(!is.na(count)) |>
    select(section_id, morphotaxon, count)

coral_contacts <- bind_rows(coral_contacts_bev, coral_contacts_niue) |>
  arrange(section_id)

coral_contacts <- coral_contacts |> 
  left_join(coral_sections |> 
              select(section_id, ps_station_id))

coral_sections <- coral_sections |> 
  left_join(coral_contacts |> 
              group_by(section_id, ps_station_id) |> 
              summarize(coral_pts = sum(count, na.rm = T))) |> 
  replace_na(list(coral_pts = 0 ))
```

```{r}
section_alignment <- lpi_sections |> 
  full_join(coral_sections |> 
              select(section_id, coral_pts),
            by = c("section_id")) |> 
  mutate(match = case_when(# Alignment
                           lpi_coral_pts > 0   & coral_pts > 0     ~ "normal",
                           lpi_coral_pts ==  0 & coral_pts  ==  0  ~ "normal",
                           # LPI data only (these become "Scleractinia")
                           lpi_coral_pts > 0   & (is.na(coral_pts) | coral_pts == 0)   ~ "lpi_only",
                           # When coral diver has pts but lpi diver does not, we turn to off transect obs
                           lpi_coral_pts == 0  & coral_pts > 0        ~ "off transect",
                           is.na(lpi_coral_pts)  & !is.na(coral_pts) & coral_pts > 0   ~ "specialist_only"),
         diff = lpi_coral_pts - coral_pts)

section_alignment |> 
  count(match, sort = T)

section_alignment |> 
  filter(match != "normal") |>
  select(section_id, lpi_pct_coral, coral_pts, match)

section_alignment <- section_alignment |> 
  filter(match != "specialist_only")

section_alignment |> 
  filter(is.na(coral_pts))
```

### Diagnostics

```{r}
library(broom)
library(patchwork)
library(glue)

# Clean data
df <- section_alignment |>
  filter(!is.na(lpi_coral_pts), !is.na(coral_pts))

# Compute stats
cor_test <- cor.test(df$lpi_coral_pts, df$coral_pts)
lm_fit   <- lm(coral_pts ~ lpi_coral_pts, data = df)
lm_glance <- glance(lm_fit)
lm_coef   <- tidy(lm_fit)

# Store key metrics
r2     <- round(lm_glance$r.squared, 3)
r      <- round(cor_test$estimate, 3)
slope  <- round(lm_coef$estimate[2], 2)
interc <- round(lm_coef$estimate[1], 2)

# Plot 1: Histogram of differences
p1 <- ggplot(df, aes(x = diff)) +
  geom_histogram(binwidth = 1, fill = "#4B9CD3", color = "white", alpha = 0.8) +
  labs(x = "Difference (LPI − Coral Points)", y = "Count") +
  theme_pristineseas()

# Plot 2: Scatter with fit and 1:1 line
p2 <- ggplot(df, aes(x = lpi_coral_pts, y = coral_pts)) +
  geom_point(color = "#E4572E", alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", color = "#4B9CD3", se = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  coord_equal() +
  labs(
    x = "LPI Coral Points",
    y = "Coral Points",
    subtitle = glue("r = {r}, R² = {r2}, slope = {slope}, intercept = {interc}")
  ) +
  theme_pristineseas()

# Combine side by side
(p1 | p2) +
  plot_annotation(
    title = "Section Alignment Diagnostics",
    subtitle = "Histogram of differences (left) and LPI vs Coral fit (right)",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5, margin = margin(b = 6)),
      plot.subtitle = element_text(size = 12, hjust = 0.5, margin = margin(b = 12))
    )
  )
```
### Apportion Coral Points

```{r}
normal_section_ids <- section_alignment$section_id[section_alignment$match == "normal"]

taxa_proportions <- coral_contacts |> 
  group_by(ps_station_id, section_id) |>
  mutate(prop_contacts = count / sum(count)) |>
  ungroup() |> 
  filter(count > 0) |> 
  select(ps_station_id, section_id, morphotaxon, prop_contacts)

match_normal <- section_alignment |> 
  filter(match == "normal") |> 
  mutate(diver = "Eric Brown") |> 
  left_join(taxa_proportions |> 
              filter(section_id %in% normal_section_ids),
            by = c("ps_station_id", "section_id"),
            suffix = c("_lpi", "_coral")) |> 
   mutate(contacts = lpi_coral_pts * prop_contacts) |> 
  filter(!is.na(ps_station_id))

match_lpi_only <- section_alignment |> 
  filter(match == "lpi_only") |> 
  mutate(morphotaxon =  "Scleractinia",
         contacts = lpi_coral_pts) |> 
  filter(!is.na(ps_station_id))

off_transect_obs <- section_alignment |>
  filter(match == "off transect") |>
    mutate(diver = "Eric Brown") |>
  left_join(taxa_proportions,
            by = c("ps_station_id", "section_id"),
            suffix = c("_lpi", "_coral")) |>
  mutate(contacts = 0,
         transect_section = "Off",
         section_id =  paste(ps_station_id, transect_section, sep = "_"))

apportioned_coral_pts <- bind_rows(match_normal, match_lpi_only) |> 
  select(-lpi_coral_pts, -lpi_pct_coral, -coral_pts, -prop_contacts, -match, -diff, -n_points) |> 
  mutate(contacts = round(contacts, 3)) |> 
  rename(count = contacts)

# We are consistnely keeping total coral points
sum(apportioned_coral_pts$count, na.rm = T)/total_lpi_coral_pts
```

```{r}
# Step 8: Combine with non-coral LPI data to create complete dataset
lpi_contacts_non_coral <- lpi_contacts |>
  filter(morphotaxon != "Hard coral") 

complete_lpi_obs <- bind_rows(lpi_contacts_non_coral,           # All non-coral categories
                               apportioned_coral_pts |> 
                                 select(-survey_type, -method, -ps_site_id, -depth_strata) 
                               ) |> 
  arrange(ps_station_id, section_id, morphotaxon) |> 
  select(section_id, everything()) |> 
  filter(count > 0 | transect_section == "Off")

complete_lpi_obs <- complete_lpi_obs |> 
  separate(transect_section, into = c("start", "end"), sep = "-", convert = TRUE, remove = FALSE) |> 
  group_by(ps_station_id, transect) |> 
  mutate(section_length    = as.numeric(end) - as.numeric(start),
         n_sections        = n_distinct(transect_section[transect_section != "Off"]),
         transect_length_m = if_else(transect_section != "Off", n_sections*section_length, 50)) |> 
  ungroup() |> 
  select(ps_station_id, exp_id, diver, depth_m, transect, transect_length_m, transect_section, morphotaxon, count)

sum(complete_lpi_obs$count, na.rm = T)/total_lpi_pts
```

```{r}
apportioned_coral_pts |> 
  group_by(ps_station_id, section_id) |>
  summarize(apportioned_total = sum(count), .groups = "drop") |> 
  left_join(lpi_sections |> 
              select(ps_station_id, section_id, lpi_coral_pts), 
            by = c("ps_station_id", "section_id")) |>
  mutate(difference = apportioned_total - lpi_coral_pts) |> 
  summarize(total_apportioned = sum(apportioned_total, na.rm = TRUE), 
            total_original    = sum(lpi_coral_pts, na.rm = TRUE),
            mean_difference   = round(mean(difference, na.rm = TRUE), 4),
            max_difference    = round(max(abs(difference), na.rm = TRUE), 4))
```

# 3. Taxonomy Resolution

```{r}
# Common Misspellings

complete_lpi_obs <- complete_lpi_obs |> 
  mutate(morphotaxon = str_replace(morphotaxon, "Lobophyton", "Lobophytum"),
         morphotaxon = str_replace(morphotaxon, "Lendenfeldeldia", "Lendenfeldia"),
         morphotaxon = str_replace(morphotaxon, "Lynbya", "Lyngbya"),
         morphotaxon = str_replace(morphotaxon, "Dendronephyta", "Dendronephthya"),
         morphotaxon = str_replace(morphotaxon, "Scleronephyta", "Scleronephthya"),
         morphotaxon = str_replace(morphotaxon, "Cymbastella", "Cymbastela"),
         morphotaxon = str_replace(morphotaxon, "Halyclonia", "Haliclona"),
         morphotaxon = str_replace(morphotaxon, "Asteriospicularia", "Asterospicularia") ,
         morphotaxon = str_replace(morphotaxon, "Coalocarteria", "Coelocarteria") ,
         morphotaxon = str_replace(morphotaxon, "Anthipathes", "Antipathes"),
         morphotaxon = str_replace(morphotaxon, "Elisella", "Ellisella"),
         morphotaxon = str_replace(morphotaxon, "hutschyanum", "kotschnyanum"),
         morphotaxon = str_replace(morphotaxon, "Microcinidae", "Microcionidae"),
         morphotaxon = str_replace(morphotaxon, "Darwillenidae", "Darwinellidae"),
         morphotaxon = str_replace(morphotaxon, "gemmacaea", "gemmacea"),
         morphotaxon = str_replace(morphotaxon, "acuelata", "aculeata"),
         morphotaxon = str_replace(morphotaxon, "Sandolitha", "Sandalolitha"),
         morphotaxon = str_replace(morphotaxon, "elephatotus", "elephantotus"),
         morphotaxon = str_replace(morphotaxon, "chaldicum|chalcidium", "chalcidicum"),
         morphotaxon = str_replace(morphotaxon, "transvera", "transversa"),
         morphotaxon = str_replace(morphotaxon, "amata", "ammata"),
         morphotaxon = str_replace(morphotaxon, "hirsutissma", "hirsutissima"),
         morphotaxon = str_replace(morphotaxon, "moniasteriata", "monasteriata"),
         morphotaxon = str_replace(morphotaxon, "Gibbsmithia", "Gibsmithia"),
         morphotaxon = str_replace(morphotaxon, "nassuta", "nasuta"),
         morphotaxon = str_replace(morphotaxon, "Milipora", "Millepora"),
         morphotaxon = str_replace(morphotaxon, "edyouxi", "eydouxi"),
         morphotaxon = str_replace(morphotaxon, "phyrgia", "phrygia"),
         morphotaxon = str_replace(morphotaxon, "urvilliana", "urvilleana"),
         morphotaxon = str_replace(morphotaxon, "Echinophyliia", "Echinophyllia"),
         morphotaxon = str_replace(morphotaxon, "Euphillia", "Euphyllia"))

complete_lpi_obs$morphotaxon[complete_lpi_obs$morphotaxon == "Favia matthai"] <- "Favia matthaii"
complete_lpi_obs$morphotaxon[complete_lpi_obs$morphotaxon == "Clavelina zobustra"] <- "Clavelina robusta"

# Summarize by mophotaxon

lpi_taxa <- complete_lpi_obs |> 
  group_by(morphotaxon) |> 
  summarise(count = round(sum(count), 2),
            .groups = "drop") |> 
  mutate(pct_count = round(100 * count / sum(count), 2)) |> 
  filter(!morphotaxon %in% c("Sand","Sediment", "Turf", "Barren", "Coral rubble", "Hard coral - dead", "Hard coral - bleached", "EAM - Epilithic Algal Matrix")) |> 
  arrange(desc(count))

# Clean

lpi_taxa <- lpi_taxa |> 
  mutate(taxon_clean = morphotaxon |>
           str_remove(regex("\\blike.*", ignore_case = TRUE)) |>          # remove "like..." and everything after
           str_remove(regex("\\bsp(p)?\\.?\\b.*", ignore_case = TRUE)) |> # remove "sp.", "spp.", etc.
           str_remove(regex("\\bcf\\.|aff\\.", ignore_case = TRUE)) |>    # remove "cf.", "aff."
           str_remove_all("[\\?\\(\\)]") |>                               # remove question marks and parens
           str_squish() |> 
           str_to_sentence()) |> 
  mutate(taxon_clean = case_when(str_detect(morphotaxon, regex("sponge", ignore_case = TRUE)) | str_detect(morphotaxon, "Clathria/Dysidea") ~ "Porifera",
                                 morphotaxon %in% c("Red algae non identified", "Red encrusting alga non identified")                 ~ "Rhodophyta",
                                 morphotaxon %in% c("Turf", "Sediment", "Coral rubble", "Barren", "EAM - Epilithic Algal Matrix")     ~ NA_character_,
                                 morphotaxon == "Soft coral" | str_detect(morphotaxon, "Gorgonia")                       ~ "Octocorallia",
                                 morphotaxon == "Scleronephthya/Stereophthya?"                                                        ~ "Nephtheidae",
                                 morphotaxon == "Tetillidae-Cynachyrella?"                                                            ~ "Tetillidae",
                                 str_detect(morphotaxon, "Hard coral")                                                                ~ "Scleractinia",
                                 morphotaxon == "Anemone unidentified"                                                                ~ "Actiniaria",
                                 morphotaxon == "Lyngbya + Halimeda"                                                                  ~ "Lyngbya",
                                 morphotaxon %in% c("CCA", "CCA undertermined")                                                         ~ "Corallinales",
                                 morphotaxon == "Halimeda mitja"                                                                      ~ "Halimeda",
                                 morphotaxon == "Halimeda taenicola mini"                                                             ~ "Halimeda taenicola",
                                 morphotaxon == "Hidrarian"                                                                           ~ "Hydrozoa",
                                 morphotaxon == "Clavelina zobustra"                                                                  ~ "Clavelina robusta",
                                 TRUE ~ taxon_clean),
         taxon_clean_bdc = bdc::bdc_clean_names(taxon_clean)$names_clean) |> 
  mutate(taxon_clean = coalesce(taxon_clean_bdc, taxon_clean),
         taxon_clean = case_when(morphotaxon == "Caulerpa serrulata f. spiralis" ~ "Caulerpa serrulata f. spiralis",
                                 morphotaxon == "Caulerpa racemosa var. peltata" ~ "Caulerpa racemosa var. peltata",
                                 morphotaxon == "Caulerpa chemnitzia varietat turbinata" ~ "Caulerpa chemnitzia var. turbinata",
                                 morphotaxon == "Halimeda lacunalis f. lata" ~ "Halimeda lacunalis f. lata",
                                 TRUE ~ taxon_clean)) |> 
  distinct(morphotaxon, taxon_clean, count, pct_count) |> 
  arrange(desc(pct_count))

lpi_taxa$taxon_clean[lpi_taxa$morphotaxon == "Favia spinosa"] <- "Favites spinosa"
lpi_taxa$taxon_clean[lpi_taxa$morphotaxon == "Montipora turgensis"] <- "Montipora turgescens"

lpi_taxa |> 
  filter(morphotaxon != taxon_clean)
```

```{r}
# Let's check against our Taxonomic Reference Table

# Create name-to-AphiaID mapping

names_to_aphiaID <- benthos_taxa_lut |> 
  select(accepted_aphia_id, taxon_name, accepted_name) |>
  pivot_longer(cols = c("taxon_name", "accepted_name"), names_to = "orig_col", values_to = "lookup_name") |>
  filter(!is.na(lookup_name)) |>
  distinct(accepted_aphia_id, lookup_name) 

lpi_taxa <- lpi_taxa |> 
  left_join(names_to_aphiaID, 
            by = c("taxon_clean" = "lookup_name")) 
  
lpi_taxa |> 
  filter(is.na(accepted_aphia_id))
```

```{r eval = F}
new_taxa <- lpi_taxa |> 
  filter(is.na(accepted_aphia_id)) |> 
  select(morphotaxon, taxon_clean, pct_count)

new_worms_records <- purrr::map_dfr(new_taxa$taxon_clean, 
                            ~worrms::wm_records_names(.x)) |> 
  select(taxon_clean = scientificname, aphia_id = AphiaID, rank, 
         name_status = status, accepted_name = valid_name, accepted_aphia_id = valid_AphiaID) |> 
  mutate(rank = str_to_lower(rank))

new_worms_records <- new_worms_records |> 
  filter(!accepted_aphia_id %in% c(145371, 377141)) 

dupes <- new_worms_records |> 
  get_dupes(taxon_clean)

deduped <- dupes |> 
  filter(name_status == "accepted") |> 
  select(-dupe_count) |> 
  distinct()

get_taxonomic_ranks <- function(id) {
  tryCatch(
    {worrms::wm_classification(id) |>
        select(rank, scientificname) |>
        pivot_wider(names_from = rank, values_from = scientificname) |>
        mutate(accepted_aphia_id = id)},
    error = function(e) {
      tibble(accepted_aphia_id = id)
    })
  }

library(furrr)

new_taxonomy_ranks <- furrr::future_map_dfr(new_worms_records$accepted_aphia_id,
                                 get_taxonomic_ranks,
                                 .options = furrr_options(seed = TRUE)) |>
  clean_names() |>
  select(accepted_aphia_id, kingdom, phylum, class, order, family, genus)

new_taxa <- new_taxa |> 
  left_join(new_worms_records) |> 
  left_join(new_taxonomy_ranks) |> 
  rename(taxon_name = taxon_clean,
         status = name_status) |> 
  select(-morphotaxon, -pct_count) 

new_taxa$phylum[new_taxa$class == "Florideophyceae" & is.na(new_taxa$phylum)] <- "Chlorophyta"
new_taxa$phylum[new_taxa$class == "Ulvophyceae" & is.na(new_taxa$phylum)] <- "Chlorophyta"
  
new_taxa <- new_taxa |> 
  mutate(functional_group = case_when(taxon_name == "Turbinaria ornata" ~ "algae_erect",
                                      order == "Scleractinia" ~ "hard_coral",
                                      order == "Corallinales" ~ "cca",
                                      class == "Cyanophyceae" ~ "cyanobacteria",
                                      genus %in% c("Halimeda", "Microdictyon", "Dictyota", "Liagora", "Yamadaella") ~ "algae_erect",
                                      genus %in% c("Dictyosphaeria", "Boodlea") ~ "turf",
                                      accepted_name == "Ceraesignum maximum" ~ "molluscs",
                                      TRUE ~ NA_character_)) 

# IUCN status

iucn_db <- read_csv(file.path(ps_paths$datasets, 
                              "iucn-redlist-marine-species",  
                              "joined_and_resolved_taxa.csv"))

iucn_priority <- c("CR", "EN", "VU", "NT", "LC", "DD", "NE")

iucn_clean <- iucn_db |> 
  rename(accepted_name = taxon_valid_name) |> 
  filter(accepted_name %in% new_taxa$accepted_name) |> 
  distinct(accepted_name, iucn_redlist_cat, iucn_taxon_id) |> 
  mutate(iucn_redlist_cat = case_when(str_detect(iucn_redlist_cat, "Critically Endangered") ~ "CR",
                                      str_detect(iucn_redlist_cat, "Endangered") ~ "EN",
                                      str_detect(iucn_redlist_cat, "Vulnerable") ~ "VU",
                                      str_detect(iucn_redlist_cat, "Near Threatened") ~ "NT",
                                      str_detect(iucn_redlist_cat, "Least Concern") ~ "LC",
                                      str_detect(iucn_redlist_cat, "Data Deficient") ~ "DD",
                                      str_detect(iucn_redlist_cat, "Extinct") ~ "EX",
                                      str_detect(iucn_redlist_cat, "Lower Risk/least concern") ~ "LC",
                                      str_detect(iucn_redlist_cat, "Lower Risk/near threatened") ~ "NT",
                                      TRUE ~ NA_character_)) |> 
  mutate(iucn_rank = match(iucn_redlist_cat, c("EX", iucn_priority))) |> 
  group_by(accepted_name) |> 
  arrange(iucn_rank) |> 
  slice(1) |> 
  ungroup() |> 
  select(-iucn_rank) |> 
  rename(iucn_cat = iucn_redlist_cat,
         iucn_sis_id = iucn_taxon_id)

new_taxa <- new_taxa |> 
  left_join(iucn_clean)

# upload to BQ

bq_table_upload(bq_table("pristine-seas", "taxa_info", "benthos"),
                values = new_taxa,
                create_disposition = "CREATE_NEVER",
                write_disposition = "WRITE_APPEND")
```

```{r}
lpi_taxa <- lpi_taxa |> 
  left_join(benthos_taxa_lut |> 
              filter(!aphia_id %in% c(224174, 1341, 224179), !is.na(accepted_aphia_id)) |> 
              distinct(accepted_aphia_id, accepted_name, rank, functional_group, family)) 
```

```{r}
lpi_taxa |> 
  mutate(rank = coalesce(rank, morphotaxon)) |> 
  group_by(rank) |> 
  summarize(n_morphotaxa = n_distinct(morphotaxon),
            n_aphia_id = n_distinct(accepted_aphia_id),
            contacts = sum(count),
            .groups = "drop") |> 
  arrange(desc(contacts)) |> 
  mutate(pct_contacts = round(100*contacts/sum(contacts), 2))

lpi_taxa |> 
  filter(!is.na(family)) |> 
  group_by(family) |> 
  summarize(n_morphotaxa = n_distinct(morphotaxon),
            n_aphia_id = n_distinct(accepted_aphia_id),
            contacts = sum(count),
            .groups = "drop") |> 
  arrange(desc(contacts)) |> 
    mutate(pct_contacts = round(100*contacts/sum(contacts), 2))

lpi_taxa |> 
  group_by(functional_group) |> 
  summarize(n_morphotaxq = n_distinct(morphotaxon),
            n_aphia_id = n_distinct(accepted_aphia_id),
            contacts = sum(count),
            .groups = "drop") |> 
  arrange(desc(contacts)) |> 
  mutate(pct_contacts = round(100*contacts/sum(contacts), 2)) # Excluding the non taxa groups e.g., barren

lpi_taxa |> filter(is.na(functional_group))
```
## Join Taxonomy to Observations

```{r}
# Combine with contacts data

lpi_observations <- complete_lpi_obs |> 
  left_join(lpi_taxa |> 
              distinct(morphotaxon, accepted_name, accepted_aphia_id, rank, functional_group, family), 
            by = "morphotaxon") |> 
  mutate(functional_group = case_when(morphotaxon %in% c("Barren", "Sediment", "Sand") ~ "sediment | rubble | barren", 
                                      morphotaxon == "Turf" ~ "turf",
                                      TRUE ~ functional_group),
         notes = NA_character_) |> 
  select(ps_station_id, exp_id, diver, depth_m, transect, transect_length_m, transect_section, 
         morphotaxon, accepted_name, accepted_aphia_id, rank, family, functional_group, count, notes) 

lpi_observations |> 
  get_dupes(ps_station_id, transect, transect_section, morphotaxon)
```

# 4. Summarize  Cover

## By Transect

```{r}
transects <- lpi_observations |> 
  filter(transect_section != "Off") |> 
  group_by(ps_station_id, transect) |> 
  summarize(divers            = paste0(sort(unique(diver),decreasing = T), collapse = " | "),
            depth_m           = round(mean(depth_m)),
            depth_strata      = stratify(depth_m),
            n_sections        = n_distinct(transect_section),
            transect_length_m = n_sections*10,
            n_points          = round(sum(count)),
            .groups = "drop") 

cover_by_transect_taxa <- lpi_observations |>
  filter(transect_section != "Off") |>
  # 1. Section-level % cover (for SD)
  group_by(ps_station_id, transect, transect_section) |>
  mutate(pct_cover_section = 100 * count / sum(count)) |>
  ungroup() |>
  # 2. Fill missing taxa × section combos
  group_by(ps_station_id, transect, diver) |>
  complete(nesting(transect_section),
           nesting(functional_group, morphotaxon, accepted_name, accepted_aphia_id, rank, family),
           fill = list(count             = 0, 
                       pct_cover_section = 0)) |>
  ungroup() |>
  # 3. Compute total contacts and SD per transect (sum across sections)
  group_by(diver, ps_station_id, transect, 
           morphotaxon, accepted_name, accepted_aphia_id, functional_group, rank, family) |>
  summarize(total_count    = sum(count),
            pct_cover_sd   = sd(pct_cover_section),
            .groups = "drop") |>
  # 4. Compute compositional % cover
  group_by(ps_station_id, transect) |>
  mutate(pct_cover = round(100 * total_count / sum(total_count), 4)) |>
  ungroup() 
```

## By Station

```{r}
lpi_stations <- transects |> 
  group_by(ps_station_id, depth_strata) |> 
  summarize(divers        = paste0(sort(unique(divers), decreasing = T), collapse = " | "),
            depth_m       = round(mean(depth_m)),
            n_transects   = n_distinct(transect),
            n_sections    = sum(n_sections),
            n_points      = sum(n_points),
            survey_dist_m = sum(transect_length_m),
            .groups = "drop") |> 
  # Add # Taxa
  left_join(lpi_observations |>
              filter(transect_section != "Off", ! morphotaxon %in% c("Barren", "Sediment", "Turf")) |> 
              group_by(ps_station_id) |>
              summarize(n_taxa = n_distinct(accepted_name[count > 0 ])),
            by = "ps_station_id") |> 
  # Add site info
  mutate(ps_site_id = str_replace(ps_station_id, "_[^_]+$", "")) |> 
  left_join(uvs_sites |> 
              distinct(exp_id, ps_site_id, region, subregion, habitat, locality, exposure),
            by = "ps_site_id") |> 
  select(ps_station_id, ps_site_id, exp_id, region, subregion, locality, habitat, exposure, 
         depth_strata, depth_m,  divers, n_transects, survey_dist_m, n_points, n_taxa)
```

```{r}
cover_by_station_taxa <- cover_by_transect_taxa |> 
  select(ps_station_id, transect, morphotaxon, accepted_name, functional_group, total_count, pct_cover, pct_cover_sd) |> 
  group_by(ps_station_id) |> 
  complete(nesting(transect), 
           nesting(morphotaxon, accepted_name, functional_group),
           fill = list(total_count  = 0,
                       pct_cover    = 0,
                       pct_cover_sd = NA_real_)) |> 
  group_by(ps_station_id, morphotaxon, accepted_name, functional_group,) |> 
  summarize(total_count  = round(sum(total_count), 3),
            pct_cover    = round(mean(pct_cover),3),
            pct_cover_sd = round(mean(pct_cover_sd),3),
            .groups = "drop") 

cover_by_station_taxa <- lpi_stations |> 
  inner_join(cover_by_station_taxa) |> 
  left_join(lpi_taxa |> 
              distinct(morphotaxon, accepted_name, accepted_aphia_id, rank, family, functional_group)) |> 
  select(ps_station_id, ps_site_id, exp_id, region, subregion, habitat, exposure, 
         depth_strata, depth_m, divers, n_transects, survey_dist_m,
         morphotaxon, accepted_name, accepted_aphia_id, rank, family, functional_group, total_count, pct_cover, pct_cover_sd)
```

```{r}
others_group <- c("forams","bryozoans","worms","echinoderms", "molluscs", "hydrozoans", "ascidians", "other_cnidarians")

grps_order <- c("cyanobacteria", "eam", "sediment | rubble | barren", "other", "turf", "algae_encrusting", "algae_erect","sponges", "soft_coral","cca", "hard_coral")

cover_by_station_group <- cover_by_station_taxa |> 
  mutate(functional_group = case_when(functional_group %in% others_group         ~ "other",
                                      functional_group %in% c("hard_coral_dead") ~ "sediment | rubble | barren",
                                      TRUE ~ functional_group)) |> 
  group_by(region, subregion, ps_site_id, ps_station_id, depth_strata, functional_group) |> 
  summarize(pct_cover = round(sum(pct_cover),2),
            .groups   = "drop") |> 
  pivot_wider(names_from = functional_group, 
              values_from = pct_cover, 
              values_fill = 0) |> 
  pivot_longer(cols = -c(ps_station_id, region, subregion, depth_strata, ps_site_id), 
               names_to = "functional_group", 
               values_to = "pct_cover") |>
  mutate(functional_group = fct_relevel(functional_group, grps_order))

cover_by_station_group_wide <- cover_by_station_group |> 
  select(ps_station_id, functional_group, pct_cover) |>
  pivot_wider(names_from = functional_group, 
              values_from = pct_cover, 
              values_fill = 0,
              names_prefix = "pct_") |> 
  janitor::clean_names() |> 
  rename(pct_algae_encrust = pct_algae_encrusting,
         pct_cyano         = pct_cyanobacteria,
         pct_coral         = pct_hard_coral,
         pct_rubble        = pct_sediment_rubble_barren)

lpi_stations <- lpi_stations |> 
  left_join(cover_by_station_group_wide)
```

# 5. QA/QC Validation

## Transect Checks

```{r}
agent_transects <- transects |> 
  create_agent(label = "LPI Transects QA/QC",
               tbl_name = "transects") |> 
  # Core fields must be complete
  rows_complete(columns = vars(ps_station_id, transect, divers, n_sections, transect_length_m, n_points),
                label = "No missing core fields",
                actions = action_levels(stop_at = 0.01)) |> 
  # Survey effort was standard
  col_vals_equal(columns = vars(n_sections),
                   value = 5,
                   label = "Five sections per transect",
                   actions = action_levels(warn_at = 0.1)) |>
  col_vals_equal(columns = vars(transect_length_m),
                   value = 50,
                   label = "Five sections per transect",
                   actions = action_levels(warn_at = 0.1)) |>
  col_vals_equal(columns = vars(n_points),
                   value = 250,
                   label = "250 points per transect",
                   actions = action_levels(warn_at = 0.1)) |>
  interrogate()

agent_transects
```

## Station Checks

```{r}
# Station QA/QC using pointblank
station_qaqc <- lpi_stations |> 
  create_agent(label = "Benthos LPI Stations QA/QC", tbl_name = "lpi_stations") |> 
  rows_distinct(ps_station_id,
                label = "Station IDs are unique",
                actions = action_levels(stop_at = 0.001)) |>
  col_vals_equal(columns = vars(n_transects), 
                 value = 1,
                 label = "Expected 1 transects per station",
                 actions = action_levels(warn_at = 0.001)) |>
  interrogate()

# Display QA/QC results
station_qaqc
```

```{r}
cover_by_station_taxa |> 
  group_by(ps_station_id) |> 
  summarize(total_cover = round(sum(pct_cover))) |> 
  filter(total_cover != 100)
```

## Observation Checks

```{r}
cover_by_station_taxa |> 
  filter(is.na(accepted_name), 
         !functional_group %in% c("sediment | rubble | barren", "turf"))
```

```{r}
lpi_observations |> 
  create_agent(label = "Benthos LPI Observations QA/QC", tbl_name = "lpi_observations") |> 
  # funcitonal groups in allowed vocab
  col_vals_in_set(columns = vars(functional_group),
                  set     = allowed_vocab$functional_groups,
                  label   = "Functional group in allowed vocabulary",
                  actions = action_levels(stop_at = 0.01)) |>
  interrogate()
```


# 6. Summary Statistics

```{r summary_stats}
#| label: tbl-summary
#| tbl-cap: "LPI survey effort by region and depth strata"

lpi_stations |> 
  group_by(region, depth_strata) |> 
  summarise(n_stations      = n(),
            mean_taxa       = round(mean(n_taxa), 1),
            mean_pct_coral    = round(mean(pct_coral), 1),
            .groups = "drop") |> 
  pivot_wider(names_from  = depth_strata,
              values_from = c(n_stations, mean_taxa, mean_pct_coral),
              values_fill = 0)  |> 
  gt(groupname_col = "region") |> 
  tab_spanner(label = "Stations", columns = starts_with("n_stations")) |> 
  tab_spanner(label = "Mean Taxa", columns = starts_with("mean_taxa")) |> 
  tab_spanner(label = "Mean % Coral", columns = starts_with("mean_pct_coral")) |> 
  cols_label(region = "region",
             n_stations_deep            = "Deep",
             n_stations_shallow         = "Shallow",
             n_stations_supershallow    = "Supershallow",
             mean_taxa_deep             = "Deep",
             mean_taxa_shallow          = "Shallow",
             mean_taxa_supershallow     = "Supershallow",
             mean_pct_coral_deep          = "Deep",
             mean_pct_coral_shallow       = "Shallow",
             mean_pct_coral_supershallow  = "Supershallow") |> 
  tab_options(table.font.size = 12, row_group.as_column = TRUE)
```

```{r}
lpi_stations |> 
  filter(exposure != "lagoon") |> 
  group_by(region) |> 
  summarise(n_stations         = n_distinct(ps_station_id),
            avg_n_taxa         = mean(n_taxa),
            across(contains("pct"), mean), 
            .groups = "drop") |> 
  mutate_if(is.numeric, round, 1)
```

```{r taxa_summary}
#| label: tbl-taxa
#| tbl-cap: "Top 10 invertebrate taxa by frequency"

taxa_summary <- cover_by_station_taxa |> 
  filter(functional_group != "sediment | rubble | barren") |> 
  mutate(accepted_name = coalesce(accepted_name, morphotaxon)) |>
  select(ps_station_id, accepted_name, total_count, pct_cover) |> 
  complete(nesting(ps_station_id),
           nesting(accepted_name),
           fill = list(pct_cover = 0,
                       total_count = 0)) |> 
  group_by(accepted_name) |> 
  summarize(total_count = round(sum(total_count)),
            freq_obs     = round(n_distinct(ps_station_id[pct_cover > 0])/length(unique(lpi_stations$ps_station_id)),3),
            mean_cover = round(mean(pct_cover), 2),
            .groups = "drop") |> 
  arrange(desc(mean_cover))

taxa_summary |> 
  slice_head(n = 10) |> 
  gt() |> 
  cols_label(accepted_name = "Scientific Name",
             total_count   = "Total Count",
             freq_obs      = "Fraction Stations Observed",
             mean_cover  = "Mean Cover (%)") |> 
  fmt_number(columns = c(mean_cover), decimals = 2) |> 
  tab_options(table.font.size = 12)
```

# 7. Map

```{r}
#| label: fig-map
#| fig-cap: "Interactive map of LPI survey stations"

# Create spatial features
lpi_sites_sf <- uvs_sites |> 
  select(ps_site_id, region, subregion, locality, habitat, exposure, latitude, longitude) |> 
  inner_join(lpi_stations |> 
               group_by(ps_site_id) |> 
               summarize(divers = paste(unique(divers), collapse = "/"),
                         strata = paste(unique(paste(depth_strata, " (", depth_m, "m)", sep = "")), 
                                        collapse = "\n"),
                         survey_dist = sum(survey_dist_m),
                         n_stations = n_distinct(ps_station_id),
                         .groups = "drop"),
             by = "ps_site_id") |> 
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 

# Create interactive map
mapview::mapview(lpi_sites_sf,
                 zcol = "exposure",
                 legend = TRUE,
                 map.types = "Esri.WorldImagery",
                 layer.name = "Exposure",
                 popup = leafpop::popupTable(lpi_sites_sf, 
                                             zcol = c("ps_site_id", "strata", "divers", "survey_dist", "habitat", "exposure"))) |> 
  leafem::addMouseCoordinates() |> 
  addFullscreenControl()
```

# 8. Exploratory Visualizations

## Site composition

```{r}
cover_by_station_group |> 
  group_by(region, ps_site_id, depth_strata, functional_group) |> 
  summarize(pct_cover = round(sum(pct_cover),2),
            .groups = "drop") |> 
  mutate(site = str_extract(ps_site_id, "[^_]+$")) |> 
  filter(depth_strata != "supershallow") |> 
  ggplot()+
  geom_col(aes(site, pct_cover, fill = functional_group))+
  facet_grid(depth_strata ~ region, scales = "free")+
  scale_fill_manual(values = ps_colors("functional_groups"))+
  labs(x = NULL, y = "Benthic composition (% of station total)",
       fill = "Functional group",
       title = "Benthic Cover by Site",
       subtitle = "Composition by depth strata")

ggsave(filename = file.path(figures_out, paste0(exp_id, "_lpi_cover_by_site.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

## Subregion composition

```{r}
cover_by_station_group |> 
  group_by(region, subregion, depth_strata, functional_group) |> 
  summarize(pct_cover = mean(pct_cover)) |> 
  ungroup() |> 
  filter(depth_strata != "supershallow") |> 
  ggplot()+
  geom_col(aes(subregion, pct_cover, fill = functional_group))+
  facet_grid(depth_strata~region, scales = "free")+
  scale_fill_manual(values = ps_colors("functional_groups"))+
  labs(x = NULL, 
       y = "Benthic composition (% of station total)",
       fill = "Functional group",
       title = "Benthic Cover by Subregion",
       subtitle = "Composition by depth strata")

ggsave(filename = file.path(figures_out, paste0(exp_id, "_lpi_cover_by_subregion.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

## Patterns by depth strata

```{r}
#| label: fig-transects-summary
#| fig-cap: "Comparison of benthic cover across depth strata showing how reef composition changes with depth"

lpi_stations |> 
  select(ps_station_id, depth_strata, pct_coral, pct_cca, pct_cyano, 
         pct_turf, pct_algae_erect, pct_rubble) |> 
  pivot_longer(cols = -c(ps_station_id, depth_strata),
               names_to = "functional_group",
               values_to = "pct_cover") |> 
  mutate(functional_group = case_when(functional_group == "pct_coral" ~ "Hard Coral",
                                      functional_group == "pct_cca" ~ "Crustose Coralline Algae",
                                      functional_group == "pct_cyano" ~ "Cyanobacteria",
                                      functional_group == "pct_turf" ~ "Turf Algae",
                                      functional_group == "pct_algae_erect" ~ "Erect Macroalgae",
                                      functional_group == "pct_rubble" ~ "Rubble",
                                      TRUE ~ functional_group),
         functional_group = factor(functional_group, 
                                   levels = c("Hard Coral", "Crustose Coralline Algae",
                                              "Turf Algae", "Erect Macroalgae", 
                                              "Cyanobacteria", "Rubble"))) |> 
  ggplot(aes(x = depth_strata, y = pct_cover, fill = depth_strata)) +
  geom_boxplot(alpha = 0.65, 
               color = "gray30", 
               outlier.shape = NA,
               linewidth = 0.5,
               show.legend = FALSE) +
  geom_jitter(width = 0.25, 
              size = 1.2, 
              alpha = 0.35, 
              color = "gray20",
              show.legend = FALSE) +
  facet_wrap(~functional_group, 
             scales = "free_y",
             ncol = 3) +
  labs(x = NULL, 
       y = "Percent Cover (%)",
       title = "Reef Benthic Composition Across Depth Strata",
       subtitle = "Distribution of functional group cover by depth strata • Boxes show median and quartiles • Points represent individual stations") +
  scale_fill_manual(values = ps_colors("depth_strata"))

ggsave(filename = file.path(figures_out, paste0(exp_id, "_lpi_patterns_by_depth.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

## % Hard coral

```{r}
library(ggbeeswarm)

lpi_stations |> 
  ggplot(aes(x = subregion, y = pct_coral)) +
  geom_quasirandom(aes(color = pct_coral), 
                   alpha = 0.4, 
                   size = 1.5,
                   width = 0.3,show.legend = F) +
  stat_summary(fun = mean, geom = "point", 
               shape = 23, size = 3, 
               fill = "black", color = "white", stroke = 1) +
  stat_summary(fun = median, geom = "crossbar",
               width = 0.5, linewidth = 0.4, color = "black") +
  facet_wrap(~region, scales = "free_x", ncol = 4) +
  labs(title = "Hard Coral Cover by Subregions",
       caption = "Each dot represents a survey station • Black diamonds show subregion averages • Gray bars indicate median values",
       x = NULL,
       y = "Hard Coral Cover (%)",
       color = "Coral Cover (%)")

ggsave(filename = file.path(figures_out, paste0(exp_id, "_lpi_coral_cover_by_subregion.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

## Most abundant taxa

```{r warning = F}
#| label: fig-top-taxa
#| fig-cap: "Percent cover distributions for the 12 most abundant taxa across depth strata, showing median and mean values with log-transformed scales to accommodate the wide range of cover values"
#| fig-width: 14
#| fig-height: 10

library(ggtext)

# Identify major taxa based on average % cover (after filling with zeros)

plot_df <- cover_by_station_taxa |> 
  select(ps_station_id, morphotaxon, accepted_name, functional_group, 
         accepted_aphia_id, rank, pct_cover, total_count) |> 
  complete(ps_station_id, 
           nesting(morphotaxon, accepted_name, functional_group, accepted_aphia_id, rank),
           fill = list(pct_cover = 0, total_count = 0))

taxa_means <- plot_df |> 
  filter(!is.na(accepted_name)) |> 
  group_by(morphotaxon, accepted_name) |> 
  summarise(mean_cover = mean(pct_cover, na.rm = TRUE),
            median_cover = median(pct_cover, na.rm = TRUE),
            total_cover = sum(pct_cover, na.rm = TRUE),
            .groups = "drop") |> 
  arrange(desc(median_cover), desc(mean_cover)) 

top_taxa <- taxa_means |> 
  head(12) |> 
  pull(morphotaxon)

# Add species names for better labels
top_taxa_names <- taxa_means |> 
  filter(morphotaxon %in% top_taxa) |> 
  mutate(# Create cleaner display names with scientific names in italics
         display_name = paste0("\n<i>", accepted_name, "</i>"),
         # Order by abundance for logical facet arrangement
         display_name = forcats::fct_reorder(display_name, median_cover, .desc = TRUE)) |> 
  select(morphotaxon, display_name)

# Compute summary stats for annotation
depth_summary <- plot_df |> 
  left_join(lpi_stations, by = "ps_station_id") |> 
  filter(morphotaxon %in% top_taxa) |> 
  group_by(morphotaxon, depth_strata) |> 
  summarise(mean_cover = mean(pct_cover, na.rm = TRUE),
            median_cover = median(pct_cover, na.rm = TRUE),
            n_obs = n(),
            .groups = "drop") |> 
  left_join(top_taxa_names, by = "morphotaxon")

# Plot

# Main plot
plot_df |> 
  left_join(lpi_stations, by = "ps_station_id") |> 
  filter(morphotaxon %in% top_taxa) |> 
  left_join(top_taxa_names, by = "morphotaxon") |> 
  ggplot(aes(x = depth_strata, y = pct_cover + 0.001)) +  # Add small constant for log scale
  # Boxplot (cleaner without fill, let color do the work)
  geom_boxplot(aes(color = morphotaxon), 
               outlier.shape = NA, 
               alpha = 0.3,
               fill = "white",
               linewidth = 0.6,
               show.legend = FALSE) +
  # Jittered points
  geom_jitter(aes(color = morphotaxon), 
              width = 0.25, 
              size = 1.5, 
              alpha = 0.5,
              show.legend = FALSE) +
  # Median as bold line segment
  geom_segment(data = depth_summary,
               aes(x = as.numeric(depth_strata) - 0.35,
                   xend = as.numeric(depth_strata) + 0.35,
                   y = median_cover + 0.001,
                   yend = median_cover + 0.001),
               color = "black",
               linewidth = 1.2,
               alpha = 0.8) +
  # Mean as diamond point
  geom_point(data = depth_summary,
             aes(y = mean_cover + 0.001),
             shape = 23,
             fill = "#e74c3c",
             color = "white",
             size = 3,
             stroke = 0.8) +
  # Facet by taxa with scientific names
  facet_wrap(~ display_name, 
             scales = "free_y",
             ncol = 4) +
  # Log scale with cleaner breaks
  scale_y_log10(breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
                labels = c("0", "0.01", "0.1", "1", "10", "100"),
                expand = expansion(mult = c(0.05, 0.15))) +
  # Labels with clear narrative
  labs(x = NULL, 
       y = "Percent Cover (%, log scale)", 
       title    = "Dominant Coral Taxa Show Distinct Depth Preferences and Cover Variability",
       subtitle = "Top 12 taxa by median abundance • Bold lines show median • Diamonds show mean • Taxa ordered by overall abundance",
       caption  = "Note: Y-axis uses log scale to visualize the wide range of cover values (0.01% to 100%)") +
  paletteer::scale_fill_paletteer_d("ggsci::default_igv") +
  paletteer::scale_color_paletteer_d("ggsci::default_igv")+
  theme(strip.text = element_markdown(size = 10, face = "bold",
                                   lineheight = 1.2,
                                   margin = margin(b = 8)))

ggsave(filename = file.path(figures_out, paste0(exp_id, "_lpi_important_taxa.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

---

# 9. Export & Upload

## Pre-upload summary

## Schema Validation

```{r validate_schema}
validate_and_export <- function(df, table_ref, label, filename) {
  
  ref_cols <- tbl(bq_connection, table_ref) |> 
    filter(!is.na(exp_id)) |> 
    head() |> 
    collect() |> 
    names()
  
  missing <- setdiff(ref_cols, names(df))
  extra <- setdiff(names(df), ref_cols)
  
  if (length(missing) > 0 || length(extra) > 0) {
    cat("\n❌ Schema mismatch for", label, "\n")
    if (length(missing)) cat("  Missing columns:", paste(missing, collapse = ", "), "\n")
    if (length(extra)) cat("  Extra columns:", paste(extra, collapse = ", "), "\n")
    stop("Fix schema before export", call. = FALSE)
  }
  
  out_path <- file.path(data_out, filename)
  df |> select(all_of(ref_cols)) |> write_csv(out_path)
  cat("✓", label, "→", out_path, "\n")
  invisible(out_path)
}
```

## CSV Exports

```{r}
stations_csv <- validate_and_export(
  df        = lpi_stations |> 
    mutate(pct_eam = NA_real_, pct_seagrass = NA_real_,pct_kelp = NA_real_, notes = NA_character_),
  table_ref = "uvs.lpi_stations",
  label     = "Stations",
  filename  = paste0(exp_id, "_uvs_lpi_stations.csv")
)

observations_csv <- validate_and_export(
  df        = lpi_observations,
  table_ref = "uvs.lpi_observations",
  label     = "Observations",
  filename  = paste0(exp_id, "_uvs_lpi_observations.csv")
)

cover_csv <- validate_and_export(
  df        = cover_by_station_taxa,
  table_ref = "uvs.lpi_cover_by_station_taxa",
  label     = "Cover by Station | Taxa",
  filename  = paste0(exp_id, "_uvs_lpi_cover_by_station_taxa.csv")
)
```

## Upload to BigQuery

```{r upload_bq, eval=FALSE}
uploads <- list(
  list(file = stations_csv,     dataset = "uvs", table = "lpi_stations"),
  list(file = observations_csv, dataset = "uvs", table = "lpi_observations"),
  list(file = cover_csv,        dataset = "uvs", table = "lpi_cover_by_station_taxa")
)

for (upload in uploads) {
  table_id <- paste0(upload$dataset, ".", upload$table)
  cat("  Uploading:", table_id, "...")
  
  bq_table_upload(
    bq_table("pristine-seas", upload$dataset, upload$table),
    values = read_csv(upload$file, show_col_types = FALSE),
    create_disposition = "CREATE_NEVER",
    write_disposition  = "WRITE_APPEND"
  )
  
  cat("✓\n")
}
```

---

# vs 2023

```{r}
all_stations <- tbl(bq_connection, "uvs.lpi_stations") |> 
  filter(exp_id %in% c("NIU_2023", "NIU_2016")) |> 
  arrange(ps_station_id) |> 
  collect()

all_cover_by_station_and_taxa <- tbl(bq_connection, "uvs.lpi_cover_by_station_taxa") |> 
  filter(exp_id %in% c("NIU_2023", "NIU_2016")) |> 
  arrange(ps_station_id) |> 
  collect()
```

```{r}
all_stations |> 
  filter(region != "Antiope", habitat == "fore_reef") |> 
  group_by(exp_id, region) |> 
  summarise(across(contains("pct"), mean), .groups = "drop") |> 
  mutate_if(is.numeric, round, 1) |> 
  arrange(region)
```

```{r}
all_cover_by_station_and_taxa |> 
  filter(region == "Beveridge") |> 
  filter(functional_group == "cca") |> 
  group_by(exp_id, accepted_name) |> 
  summarize(sum(total_count))
```



