---
title: "Benthos LPI - Data Pipeline"
date: today
format: 
  html:
    theme: minty
    self-contained: true
    code-fold: true
    toc: true 
    toc-depth: 3
    toc-location: right
    html-table-processing: none
execute:
  fig-width: 10
---

```{r setup, message = F, warning = F, fig.width = 10, fig.height = 10, echo = F}
options(scipen = 999)

# Hook to format inline numeric expressions with comma separators:
knitr::knit_hooks$set(inline = function(x) {
  # For non-numeric values, just return as character
  if (!is.numeric(x)) {return(as.character(x))}
  # Format numbers with comma as big.mark, no scientific notation
  else format(x, big.mark = ",", scientific = FALSE)})

library(sf)
library(hms)
library(readxl)
library(janitor)
library(lubridate)
library(gt)
library(pointblank)
library(tidyverse)
library(bigrquery)
library(leaflet)
library(leaflet.extras)
library(ggtext)
library(furrr)
library(pristineseas)

# ---- Parameters -----------------------------------------------------------
exp_id   <- "NIU_2016"

ps_paths <- pristineseas::get_drive_paths()

prj_path <- file.path(ps_paths$projects, "legacy-db")

# ---- Auth & Connection ----------------------------------------------------
bigrquery::bq_auth(email = "marine.data.science@ngs.org")
bq_connection <- DBI::dbConnect(bigrquery::bigquery(), project = "pristine-seas")

set_theme(pristineseas::theme_pristineseas())

data_out <- file.path(prj_path, "data/output/uvs")

figures_out <- file.path(prj_path, "figures/uvs")

legacy_path <- file.path(ps_paths$science, "legacy data/Niue/")
```

# Overview

This documentation outlines the end-to-end pipeline for processing fish Belt Line Transect (BLT) survey data collected during Pristine Seas expeditions. The pipeline ingests raw field data, standardizes formats, performs taxonomy lookups, applies rigorous quality assurance/quality control (QA/QC), computes station-level summaries, and prepares validated outputs for the Pristine Seas Science Database in BigQuery.

# 1. Data Ingestion

## Reference Tables

We begin by reading the UVS (Underwater Visual Survey) site information already processed.

```{r load_references}
uvs_sites <- tbl(bq_connection, "uvs.sites") |> 
   filter(exp_id == !! exp_id) |> 
   collect()

# Validate site data
cat("Sites loaded:", nrow(uvs_sites), "\n")
cat("Regions:",      unique(uvs_sites$region), "\n")
cat("Subregions:",   unique(uvs_sites$subregion), "\n")

# Fish taxonomy reference
fish_taxa_lut <- tbl(bq_connection, "taxa_info.fish") |> 
  filter(!is.na(accepted_name)) |> 
  collect()

cat("✓ Loaded", nrow(fish_taxa_lut), "fish taxa\n")
```

## Fieldbooks

```{r}
raw_obs <- read_xlsx(file.path(legacy_path, "FISH","Niue_fish_data_Pristine_Seas.xlsx")) |> 
  clean_names() |> 
   rename(station_label = depth_strata,
         count          = number,
         subregion      = reef,
         site           = station,
         taxon     = name_fishbase,
         datatype       = datatyp,
         min_tl_cm = min,
         max_tl_cm = max) |> 
  mutate(exp_id         = exp_id,
         ps_site_id     = paste(exp_id, "uvs",str_pad(site, 3, pad = "0"), sep = "_"),
         region         = "Niue",
         subregion      = str_to_title(subregion),
         station_label  = str_to_lower(station_label),
         terminal_phase = NA,
         obs_id            = paste("NIU_2016_fish", diver, str_pad(row_number(), 3, pad = "0"), sep = "_"),
         transect_length_m = 25,
         diver = case_when(diver == "AMF" ~ "Alan Friedlander",
                           diver == "JEC" ~ "Jenn Caselle",
                           diver == "PB"  ~ "Pauline Bosserelle")) |> 
  filter(datatype != "QUAL") |> 
  filter(!is.na(count)) |> 
  select(obs_id, exp_id, ps_site_id, diver, depth_m, station_label, transect, transect_length_m, taxon , min_tl_cm, max_tl_cm, count, terminal_phase)


raw_obs <- raw_obs |> 
  # convert JC depths
  mutate(depth_m = if_else(diver == "Jenn Caselle" & station_label != "backreef", round(depth_m*0.3048), depth_m))
```

# 2. Data Transformation

```{r}
# --- Calculate average total length and assign transect width -----------------
# This step validates length bins, ensures no logical inconsistencies, and 
# defines the appropriate transect width (2 m or 4 m) based on fish size.

## Validate input lengths ------------------------------------------------------

bad_entries <- raw_obs |> 
  filter(min_tl_cm > max_tl_cm)

if (nrow(bad_entries) > 0) {
  message(cli::col_red(paste0("⚠️  Found ", nrow(bad_entries), " entries where min_tl_cm > max_tl_cm")))
  print(bad_entries)
  stop(cli::col_red("Please correct these entries before proceeding."))
} else {
  message(cli::col_green("✓ No length inconsistencies found."))
}

## Compute mean total length and transect width --------------------------------
raw_obs <- raw_obs |> 
  mutate(length_cm = rowMeans(across(c(min_tl_cm, max_tl_cm)), na.rm = TRUE),
         transect_width_m = if_else(length_cm >= 20, 4, 2)) |> 
  select(-min_tl_cm, -max_tl_cm) |> 
  relocate(length_cm, .after = count) |> 
  relocate(transect_width_m, .after = transect_length_m)
```

```{r}
# --- Derive transect- and station-level metadata ------------------------------
# Each transect represents a distinct belt within a site. Transect area incorporates both small-fish (2 m) and large-fish (4 m) belts. 

## Aggregate transects ---------------------------------------------------------
transects <- raw_obs |> 
  group_by(ps_site_id, diver, station_label, transect, transect_length_m) |> 
  summarise(avg_depth_m = round(mean(depth_m, na.rm = TRUE), 0),
            .groups = "drop") |> 
  mutate(transect_area_m2 = transect_length_m * (4 + 2))  # 4m (large fish) + 2m (small fish)

## Aggregate stations ----------------------------------------------------------

stations <- transects |> 
  group_by(ps_site_id, station_label) |>
  summarise(divers         = paste0(sort(unique(diver),decreasing = T), collapse = " | "),
            avg_depth_m    = round(mean(avg_depth_m, na.rm = TRUE), 0),
            n_transects    = n_distinct(transect),
            survey_dist_m  = n_transects * unique(transect_length_m),
            survey_area_m2 = survey_dist_m * (4 + 2),
             .groups = "drop") %>% 
  # Calculate average depth, stratify, and create station IDs
  mutate(depth_strata  = stratify(avg_depth_m),
         ps_station_id = paste0(ps_site_id, "_", station_suffix(depth_strata))) |>
  select(ps_station_id, ps_site_id, depth_m = avg_depth_m, 
         depth_strata, divers, station_label, n_transects, survey_dist_m, survey_area_m2)

# Assign station IDs and strata

raw_obs <- raw_obs |> 
  left_join(stations |> 
              select(ps_station_id, ps_site_id, station_label, depth_strata),
            by = c("ps_site_id",  "station_label")) |> 
  relocate(ps_station_id, .after = ps_site_id) |> 
  relocate(depth_strata, .after = ps_station_id)
```

# 3. Taxonomy Resolution

Match observed species with the reference taxonomy database and identify any issues (observed lengths > max lengths).

```{r}
fish_taxa <- raw_obs |> 
  mutate(taxon = bdc::bdc_clean_names(taxon)$names_clean) |> 
  count(taxon, sort = T)

# Create name-to-AphiaID mapping

names_to_aphiaID <- fish_taxa_lut |> 
  select(accepted_aphia_id, taxon_name, accepted_name) |>
  pivot_longer(cols = c("taxon_name", "accepted_name"), names_to = "orig_col", values_to = "lookup_name") |>
  filter(!is.na(lookup_name)) |>
  distinct(accepted_aphia_id, lookup_name) 

fish_taxa <- fish_taxa |> 
  left_join(names_to_aphiaID, 
            by = c("taxon" = "lookup_name")) 
```

## New taxa

```{r}
# New taxa not in reference database
new_taxa <- fish_taxa |> 
  filter(is.na(accepted_aphia_id), !is.na(taxon)) |> 
  distinct(taxon)

if (nrow(new_taxa) > 0) {
  message(cli::col_red(paste0("⚠️  Found ", nrow(new_taxa), " taxa not in our DB")))
  print(new_taxa)
  stop(cli::col_red("Let's add them before proceeding."))
} else {
  message(cli::col_green("✓ All taxa in our DB."))
}
```

```{r, eval = F}
# 1. Get WoRMS and FishBase identifiers
# ----------------------------------------------------------------------------

worms_raw <- worrms::wm_records_names(new_taxa$taxon) |> 
  bind_rows() |> 
  mutate(rank = str_to_lower(rank)) |> 
  select(taxon_name       = scientificname, 
         rank,
         aphia_id          = AphiaID, 
         status, 
         accepted_name     = valid_name, 
         accepted_aphia_id = valid_AphiaID) 

fb_taxa <- rfishbase::load_taxa() |> 
  distinct(Species, fb_spec_code = SpecCode)

new_taxa <- worms_raw |> 
  left_join(fb_taxa, by = c("accepted_name" = "Species"))

# 2. Get taxonomic hierarchy from WoRMS
# ----------------------------------------------------------------------------

get_taxonomic_ranks <- function(id) {
  tryCatch(
    {
      worrms::wm_classification(id) |>
        select(rank, scientificname) |>
        pivot_wider(names_from = rank, values_from = scientificname) |>
        mutate(accepted_aphia_id = id)
    },
    error = function(e) tibble(accepted_aphia_id = id)
  )
}

library(furrr)

taxonomy_ranks <- new_taxa |> 
  distinct(accepted_aphia_id) |> 
  filter(!is.na(accepted_aphia_id)) |> 
  pull(accepted_aphia_id) |> 
  future_map(get_taxonomic_ranks, .options = furrr_options(seed = TRUE), .progress = TRUE) |> 
  list_rbind() |> 
  clean_names() |> 
  select(accepted_aphia_id, kingdom, phylum, class, order, family, genus)

new_taxa <- new_taxa |> 
  left_join(taxonomy_ranks, by = "accepted_aphia_id")

# 3. Get common names
# ----------------------------------------------------------------------------

fb_species_db <- rfishbase::species()

common_names <- fb_species_db |> 
  distinct(fb_spec_code = SpecCode, common_name = FBname) |> 
  filter(!is.na(common_name))

common_fams <- read_csv(file.path(ps_paths$projects, "legacy-db", "data/raw/taxonomy/fish/fish_common_families.csv"),
                        show_col_types = FALSE) |> 
  clean_names()

new_taxa <- new_taxa |> 
  left_join(common_fams, by = "family") |> 
  left_join(common_names, by = "fb_spec_code")

# 4. Get trophic traits
# ----------------------------------------------------------------------------

trophic_guilds <- read_csv(file.path(ps_paths$projects, "legacy-db", "data/raw/taxonomy/fish/Parravicini_et_al_pbio.3000702.s004.csv")) |> 
  clean_names() |> 
  mutate(taxon_name = str_replace_all(species, "_", " ")) |> 
  select(family, taxon_name, trophic_guild = trophic_guild_predicted_text) |> 
  mutate(trophic_guild = str_to_lower(trophic_guild) |> 
           str_replace_all("-", "_") |> 
           str_replace_all(" ", "_") |> 
           str_replace_all("/", " | "))

akiona_df <- readxl::read_excel(file.path(ps_paths$projects, "legacy-db", "data/raw/taxonomy/fish/Akiona_et_al_bb29484643_1_1.xlsx"),
                                sheet = "Length-Weight Parameter Table") |> 
  clean_names() |> 
  mutate(taxon = str_replace_all(taxon, "Plectorhincus", "Plectorhinchus"),
         taxon = str_replace_all(taxon, "nagasakensis", "nagasakiensis"),
         taxon = case_when(taxon == "Pristiapogon trimaculatus" ~ "Pristicon trimaculatus",
                           TRUE ~ taxon)) |> 
  mutate(taxon = str_squish(taxon),
         taxon = str_remove_all(taxon, "\\bspecies\\b"),
         taxon = str_trim(taxon)) |> 
  filter(!str_detect(taxon, " x ")) |> 
  distinct(taxon, trophic_group = trophic_classification, lw_a = a,  lw_b = b,  ltl_ratio = source_length_to_total_length_ratio, 
           lw_type = source_length_type) |> 
  mutate(trophic_group = str_to_lower(trophic_group) |> 
           str_replace("-", "_") |> 
           str_replace("/", " | "))

new_taxa <- new_taxa |> 
  left_join(trophic_guilds |> 
              distinct(taxon_name, trophic_guild), by = c("accepted_name" = "taxon_name")) |> 
  left_join(akiona_df |> 
              distinct(taxon, trophic_group),
            by = c("accepted_name" = "taxon")) |> 
  mutate(trophic_group = "top_predator")

# 5. Get length-weight parameters

new_taxa <- new_taxa |> 
  left_join(akiona_df |> 
              mutate(lw_source = "Akiona et al. (2025)") |> 
              select(accepted_name = taxon, lw_a, lw_b, ltl_ratio, lw_type, lw_source),
            by = "accepted_name")

if (nrow(new_taxa |> filter(is.na(lw_a))) > 0) {
  message(cli::col_red(paste0("⚠️  Found ", nrow(new_taxa |> filter(is.na(lw_a))), " taxa without length-weight parameters")))
} else {
  message(cli::col_green("✓ All new taxa have length-weight parameters."))
}

fb_estimates <- rfishbase::estimate() |> 
  clean_names() |> 
  mutate(tl_max_source = "FishBase",
         lw_source     = "FishBase",
         ltl_ratio     = 1,
         lw_type       = "TL") |> 
  select(fb_spec_code  = spec_code,
         tl_max        = max_length_tl,
         lw_a          = a,
         lw_b          = b,
         lw_source, 
         tl_max_source,
         ltl_ratio,
         lw_type)

new_taxa <- new_taxa |> 
  select(-lw_a, -lw_b, -ltl_ratio, -lw_source, -lw_type) |>
  left_join(fb_estimates |> 
              select(fb_spec_code, lw_a, lw_b, ltl_ratio, lw_type, lw_source, tl_max, tl_max_source),
            by = "fb_spec_code") 

# 6. Get habitat and fishery info
# ----------------------------------------------------------------------------

new_taxa <- new_taxa |> 
  left_join(fb_species_db |> 
              select(fb_spec_code = SpecCode, habitat_zone = DemersPelag, fishery_importance = Importance),
            by = "fb_spec_code")

# 7. Get IUCN Red List status
# ----------------------------------------------------------------------------

iucn_db <- read_csv(file.path(ps_paths$datasets, "iucn-redlist-marine-species", "joined_and_resolved_taxa.csv"),
                    show_col_types = FALSE)

iucn_priority <- c("CR", "EN", "VU", "NT", "LC", "DD", "NE")

iucn_clean <- iucn_db |> 
  rename(accepted_name = taxon_valid_name) |> 
  filter(accepted_name %in% fish_taxa$accepted_name) |> 
  distinct(accepted_name, iucn_redlist_cat, iucn_taxon_id) |> 
  mutate(iucn_redlist_cat = case_when(str_detect(iucn_redlist_cat, "Critically Endangered")      ~ "CR",
                                      str_detect(iucn_redlist_cat, "Endangered")                 ~ "EN",
                                      str_detect(iucn_redlist_cat, "Vulnerable")                 ~ "VU",
                                      str_detect(iucn_redlist_cat, "Near Threatened")            ~ "NT",
                                      str_detect(iucn_redlist_cat, "Least Concern")              ~ "LC",
                                      str_detect(iucn_redlist_cat, "Data Deficient")             ~ "DD",
                                      str_detect(iucn_redlist_cat, "Extinct")                    ~ "EX",
                                      str_detect(iucn_redlist_cat, "Lower Risk/least concern")   ~ "LC",
                                      str_detect(iucn_redlist_cat, "Lower Risk/near threatened") ~ "NT",
                                      TRUE ~ NA_character_)) |> 
  mutate(iucn_rank = match(iucn_redlist_cat, c("EX", iucn_priority))) |> 
  group_by(accepted_name) |> 
  arrange(iucn_rank) |> 
  slice(1) |> 
  ungroup() |> 
  select(-iucn_rank) |> 
  rename(iucn_cat = iucn_redlist_cat,
         iucn_sis_id = iucn_taxon_id)

new_taxa <- new_taxa |> 
  left_join(iucn_clean) 

# 8. Upload to BigQuery
# ----------------------------------------------------------------------------

bq_table_upload(bq_table("pristine-seas", "taxa_info", "fish"),
                values             = new_taxa,
                create_disposition = "CREATE_NEVER",
                write_disposition  = "WRITE_APPEND")
```

## Add traits

```{r}
# Join with taxonomy information

fish_taxa <- fish_taxa |> 
  left_join(fish_taxa_lut |> 
              distinct(accepted_aphia_id, accepted_name, rank, tl_max, lw_a, lw_b, ltl_ratio, trophic_group, family),
            by = c("accepted_aphia_id" = "accepted_aphia_id"))
```

### Missing  params

```{r}
# Taxa without length-weight parameters
missing_params <- fish_taxa |> 
  filter(is.na(lw_a), !is.na(accepted_aphia_id))

if (nrow(missing_params) > 0) {
  message(cli::col_red(paste0("⚠️  Found ", nrow(missing_params), "taxa without length-weight parameters:\n")))
} else {
  message(cli::col_green("✓ All taxa have length-weight parameters."))
  }
```

### Missing Trophic group

```{r}
#Taxa without trophic classification
missing_trophic <- fish_taxa |> 
  filter(is.na(trophic_group), !is.na(accepted_aphia_id))

if (nrow(missing_trophic) > 0) {
  message(cli::col_red(paste0("⚠️  Found ", nrow(missing_trophic), " taxa without trophic group")))
  missing_trophic |> 
    select(taxon, accepted_name, family) 
  } else {
  message(cli::col_green("✓ All taxa have trophic group"))
  }

fish_taxa$trophic_group[fish_taxa$accepted_name == "Synodus rubromarmoratus"] <- "top_predator"
```

## Join Taxa info to Observations

```{r}
clean_obs <- raw_obs |> 
  mutate(taxon = bdc::bdc_clean_names(taxon)$names_clean) |> 
  left_join(fish_taxa, 
            by = "taxon") |> 
  select(obs_id, ps_station_id, exp_id,
         diver, depth_m, station_label, transect, transect_length_m, transect_width_m,
         accepted_name, taxon, accepted_aphia_id, rank, family, trophic_group,
         count, length_cm, lw_a, lw_b, ltl_ratio, tl_max, terminal_phase)
```

# 4.  QA/QC Validation

### Observations

```{r}
# Observation QA/QC

create_agent(tbl = clean_obs, 
             label = "Fish BLT Observations QA/QC", 
             tbl_name = "raw_obs") |> 
  col_vals_not_null(columns = vars(ps_station_id, transect, accepted_name, count),
                    label = "Core fields are not null",
                    actions = action_levels(warn_at = 0.001)) |>
  col_vals_lte(columns = vars(length_cm),
               value   = vars(tl_max),
               label = "Length under max reported",
               actions = action_levels(warn_at = 0.001), 
               na_pass = TRUE) |>
  col_vals_gte(columns = vars(count),
               value = 1,
               label = "Count is positive",
               actions = action_levels(stop_at = 0.001)) |>
  interrogate()
```

```{r}
length_anomalies <- clean_obs |>
  filter(length_cm > tl_max) |>
  mutate(abs_diff = length_cm - tl_max,
         pct_diff = round(100 * abs_diff / tl_max)) |>
  select(obs_id, accepted_name, 
         diver, transect, count, length_cm,tl_max, pct_diff, abs_diff) |> 
  arrange(desc(pct_diff))

# Apply correction heuristics - Cap at max TL if difference > 5cm

clean_obs <- clean_obs |> 
  mutate(abs_diff  = if_else(!is.na(tl_max) & length_cm > tl_max, 
                             length_cm - tl_max, 
                             NA_real_),
         pct_diff  = if_else(!is.na(abs_diff), 
                             round(100 * abs_diff / tl_max), 
                             NA_real_),
         length_cm = case_when(!is.na(tl_max) & length_cm > tl_max & abs_diff >= 5 ~ round(tl_max), 
                               TRUE  ~ length_cm)) |> 
    select(obs_id, ps_station_id, exp_id,
         diver, depth_m, station_label, transect, transect_length_m, transect_width_m,
         accepted_name, accepted_aphia_id, rank, family, trophic_group,
         count, length_cm, lw_a, lw_b, ltl_ratio, tl_max, terminal_phase)
```

### Trasects

```{r}
transects |> 
  create_agent(label = "Fish BLT Tranects QA/QC", tbl_name = "transects") |> 
  # Core fields not null
  col_vals_not_null(columns = vars(ps_site_id, diver, station_label, transect),
                    label = "Core fields are not null",
                    actions = action_levels(warn_at = 0.001)) |> 
  # Depth is less than 30 
  col_vals_lte(columns = vars(avg_depth_m),
               value   = 30,
               label = "Depth is less than 30 m",
               actions = action_levels(warn_at = 0.001)) |>
  # Transect length is 25
  col_vals_equal(columns = vars(transect_length_m), 
                 value = 25,
                 label = "Standard Transect length",
                 actions = action_levels(warn_at = 0.001)) |>
  interrogate()
```


### Stations

```{r}
stations |> 
  create_agent(label = "Fish BLT Stations QA/QC", tbl_name = "raw_stations") |> 
  # Core fields not null
  col_vals_not_null(columns = vars(ps_station_id, ps_site_id, depth_m, depth_strata, n_transects),
                    label = "Core fields are not null",
                    actions = action_levels(warn_at = 0.001)) |>
  # Unique station IDs
  rows_distinct(ps_station_id,
                label = "Station IDs are unique",
                actions = action_levels(warn_at = 0.001)) |> 
  # Actual depth strata matches field label
  col_vals_equal(columns = vars(station_label), 
                 value = vars(depth_strata),
                 label = "Station label matches depth strata",
                 actions = action_levels(warn_at = 0.001)) |>
  col_vals_equal(columns = vars(n_transects), 
                 value = 3,
                 label = "3 transects per station",
                 actions = action_levels(warn_at = 0.001)) |>
  interrogate()
```

```{r}
# Create final station dataframe and join with site metadata

stations <- stations |> 
  select(-station_label) |> 
  left_join(uvs_sites |> 
              select(ps_site_id, region, exp_id, subregion, locality, habitat, exposure),
            by = "ps_site_id") |> 
  select(ps_station_id, ps_site_id, exp_id,
         region, subregion, locality, habitat, exposure,
         depth_strata, depth_m, divers,
         n_transects, survey_dist_m, survey_area_m2) 
```

# 5. Calculate Biomass

## Biomass by station and taxa

Convert length and count observations to biomass metrics using species-specific length-weight parameters.

  - Individual biomass is calculated using length-weight parameters from `taxa_info.fish`.
  - Transect and station-level summaries are generated to quantify total biomass per taxa.
  
Here, a critical step takes place. To complete the dataset and ensure that all taxa are represented at the station level, we use the `complete` function. This function fills in any missing combinations of transects and taxa with zeros for the biomass and count fields. This is important for ensuring that our analysis captures all taxa present at each station, even if they were not observed in every transect
  
```{r}
# Per-observation biomass

biomass_per_obs <- clean_obs |> 
  mutate(# Biomass calculations
         avg_biomass   = lw_a * (length_cm * ltl_ratio)^lw_b,
         biomass_g     = round(count * avg_biomass, 4), # round to 4 digits
         ) |> 
  select(-tl_max, -lw_a, -lw_b, -ltl_ratio, -avg_biomass)

# Transect x taxa aggregation

biomass_by_transect_and_taxa <- biomass_per_obs |> 
  mutate(# Size-dependent transect area
         density_m2    = count / (transect_length_m*transect_width_m),
         biomass_gm2   = round(biomass_g / (transect_length_m*transect_width_m), 4)) |> 
  group_by(ps_station_id, diver, station_label, transect, 
           accepted_name, accepted_aphia_id, rank, family, trophic_group) |> 
  summarize(sum_length_count = sum(length_cm * count, na.rm = TRUE),
            min_length = min(length_cm, na.rm = TRUE),
            max_length = max(length_cm, na.rm = TRUE),
            across(c(count, biomass_g, biomass_gm2, density_m2), 
                   \(x) sum(x, na.rm = TRUE)),
            .groups = "drop")

# Complete missing taxa-transect combos within region

biomass_by_station_and_taxa <- biomass_by_transect_and_taxa |>
  left_join(stations |> 
              distinct(ps_station_id, region),
            by = "ps_station_id") |>
  group_by(region) |> 
  complete(nesting(ps_station_id, diver, station_label, transect),
           nesting(accepted_name, accepted_aphia_id, rank, family, trophic_group),
           fill = list(sum_length_count = NA,
                       min_length       = NA,
                       max_length       = NA,
                       count            = 0,
                       biomass_g        = 0,
                       density_m2       = 0,
                       biomass_gm2      = 0)) |> 
  group_by(ps_station_id, accepted_name, accepted_aphia_id, rank, family, trophic_group) |>
  summarise(total_count       = sum(count, na.rm = TRUE),
            avg_count         = round(mean(count),2),
            avg_length_cm     = round(if_else(total_count > 0, sum(sum_length_count, na.rm = TRUE) / sum(total_count, na.rm = TRUE), NA_real_),2),
            min_length_cm     = if_else(total_count > 0, 
                                        if (all(is.na(min_length))) NA_real_ else min(min_length, na.rm = TRUE), 
                                        NA_real_),
            max_length_cm     = if_else(total_count > 0, 
                                        if (all(is.na(max_length))) NA_real_ else max(max_length, na.rm = TRUE), 
                                        NA_real_),
            avg_density_m2        = round(mean(density_m2), 5),
            avg_biomass_gm2       = round(mean(biomass_gm2),5),
            .groups = "drop")

biomass_by_station_and_taxa <- biomass_by_station_and_taxa |>
  filter(total_count > 0) |> 
  left_join(stations |> 
              select(ps_station_id, ps_site_id, exp_id, region, subregion, habitat, exposure, depth_strata, depth_m, divers, n_transects, survey_area_m2),
            by = "ps_station_id") |> 
  select(ps_station_id, ps_site_id, exp_id, 
         region, subregion, habitat, exposure, depth_strata, depth_m, divers, n_transects, survey_area_m2,
         accepted_name, accepted_aphia_id, rank, family, trophic_group,
         total_count, avg_count, avg_length_cm, min_length_cm, max_length_cm,
         avg_density_m2, avg_biomass_gm2)
```

```{r}
biomass_by_station_and_group <- biomass_by_station_and_taxa |> 
  group_by(ps_station_id, trophic_group) |> 
  summarize(avg_biomass_gm2 = round(sum(avg_biomass_gm2),2),
            .groups = "drop") |> 
  pivot_wider(names_from = trophic_group, values_from = avg_biomass_gm2, values_fill = 0) |> 
  pivot_longer(cols = -c(ps_station_id), names_to = "trophic_group", values_to = "avg_biomass_gm2") %>% 
  group_by(ps_station_id) %>% 
  mutate(pct_biomass = round(100 * avg_biomass_gm2 / sum(avg_biomass_gm2), 1)) |> 
  ungroup()
```

# 6. Summary Statistics

## Station Summary

Aggregate metrics at the station level 

```{r}
# Transect-level summaries

transect_summary <- biomass_by_transect_and_taxa |> 
  group_by(ps_station_id, diver, station_label, transect) |> 
  summarize(n_taxa      = n_distinct(accepted_aphia_id[count>0]), 
            total_count = sum(count),
            density_m2  = sum(density_m2),
            biomass_gm2 = sum(biomass_gm2),
            .groups = "drop") 

# Average taxa richness per transect

avg_taxa_per_transect <- transect_summary |> 
  group_by(ps_station_id) |>
  summarise(avg_taxa = round(mean(n_taxa), 2),
            .groups = "drop") 

# Station-level aggregation

station_stats <- biomass_by_station_and_taxa |> 
  group_by(ps_station_id) |> 
  summarize(n_taxa          = n_distinct(accepted_aphia_id[total_count > 0]),
            total_count     = sum(total_count),
            avg_count       = round(sum(avg_count), 2),
            avg_density_m2  = round(sum(avg_density_m2),3),
            avg_biomass_gm2 = round(sum(avg_biomass_gm2),3),
            .groups = "drop")

# % biomass by trophic group

biomass_pcts_by_station <- biomass_by_station_and_group %>%  
  select(-avg_biomass_gm2) %>% 
  pivot_wider(names_from = trophic_group, values_from = pct_biomass, names_prefix = "pct_", values_fill = 0) |> 
  janitor::clean_names() |> 
  rename(pct_herbi_detri = pct_herbivore_detritivore,
         pct_lower_carni = pct_lower_carnivore,
         pct_plankti     = pct_planktivore)

# Final station summary

station_summary <- stations |> 
  left_join(station_stats, by = "ps_station_id") |> 
  left_join(avg_taxa_per_transect, by = "ps_station_id") |> 
  left_join(biomass_pcts_by_station, by = "ps_station_id")

station_summary <- station_summary |> 
  mutate(notes = NA_character_) 
```

```{r}
station_summary |> 
  #filter(exposure != "lagoon") |> 
  group_by(subregion) |> 
  summarise(n_stations         = n_distinct(ps_station_id),
            avg_n_taxa         = mean(n_taxa),
            mean_biomass_gm2   = mean(avg_biomass_gm2),
            median_biomass_gm2 = median(avg_biomass_gm2),
            across(contains("pct"), mean), 
            .groups = "drop") |> 
  group_by(subregion) |> 
  summarise(across(where(is.numeric), mean),
            .groups = "drop") |>
  mutate_if(is.numeric, round, 1)
```

# 7. Map

Now, lets visualize the distribution of our sampling effort and summarize it by region and subregion to inspect any potential outliers.

```{r map}
#| label: fig-map
#| fig-cap: "Interactive map of BLT survey stations"

# Create spatial features

fish_sites_sf <- uvs_sites |> 
  select(ps_site_id, region, subregion, locality, habitat, exposure, latitude, longitude) |> 
  inner_join(
    stations |> 
      group_by(ps_site_id) |> 
      summarize(team = paste(unique(divers), collapse = "/"),
                strata = paste(unique(paste(depth_strata, " (", depth_m, "m)", sep = "")), 
                               collapse = "\n"),
                survey_area = sum(survey_area_m2),
                n_stations = n_distinct(ps_station_id),
                .groups = "drop"),
    by = "ps_site_id") |> 
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 

# Create interactive map
mapview::mapview(fish_sites_sf,
                 zcol = "exposure",
                 legend = TRUE,
                 map.types = "Esri.WorldImagery",
                 layer.name = "Exposure",
                 popup = leafpop::popupTable(fish_sites_sf, 
                                             zcol = c("ps_site_id", "strata", "team", "survey_area", "habitat", "exposure"))) |> 
  leafem::addMouseCoordinates() |> 
  addFullscreenControl()
```

# 8. Exploratory Visualizations

### missing data 

```{r}
#| label: fig-missingness-stations
#| fig-cap: "Missingness plot for station summary"
#| 
naniar::vis_miss(station_summary)
```

### by site

```{r}
biomass_by_station_and_group |> 
  left_join(stations %>% 
              select(ps_site_id, depth_strata, region, ps_station_id), by = "ps_station_id") %>% 
  group_by(region, ps_site_id, depth_strata, trophic_group) |> 
  summarize(avg_biomass_gm2 = round(sum(avg_biomass_gm2),2),
            .groups = "drop") |> 
  mutate(site = str_extract(ps_site_id, "[^_]+$")) |> 
  mutate(trophic_group = factor(trophic_group, 
                                 levels = rev(c("shark", "top_predator", "lower_carnivore", "herbivore | detritivore", "planktivore" )))) |> 
  #filter(depth_strata != "supershallow") |> 
  ggplot()+
  geom_col(aes(site, avg_biomass_gm2, fill = trophic_group))+
  facet_grid(depth_strata ~ region, scales = "free")+
  scale_fill_manual(values = ps_colors("trophic_group"))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  labs(x = NULL, y = "Fish biomass (g/m²)",
       fill = "",
       title = "Fish biomass composition by site, region, and depth",
       subtitle = exp_id)

ggsave(filename = file.path(figures_out, paste0(exp_id, "_fish_biomass_by_site.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

### by subregion

```{r}
biomass_by_station_and_group |> 
  mutate(trophic_group = factor(trophic_group, 
                                 levels = rev(c("shark", "top_predator", "lower_carnivore", "herbivore | detritivore", "planktivore" )))) |> 
  left_join(stations %>% 
              select(ps_site_id, depth_strata, region, subregion, ps_station_id), 
            by = "ps_station_id") %>% 
  group_by(region, subregion, depth_strata, trophic_group) |> 
  summarize(avg_biomass_gm2 = mean(avg_biomass_gm2)) |> 
  ungroup() |> 
  ggplot()+
  geom_col(aes(subregion, avg_biomass_gm2, fill = trophic_group))+
  facet_grid(depth_strata~region, scales = "free")+
  scale_fill_manual(values = ps_colors("trophic_group"))+
  labs(x = NULL, y = "Fish biomass (g/m²)",
       fill = "",
       subtitle = exp_id,
       title = "Fish biomass composition by subregion, and depth")+
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"),
        legend.position = "bottom",
        legend.key.height = unit(4, "mm"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(filename = file.path(figures_out, paste0(exp_id, "_fish_biomass_by_subregion.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

## by strata

Examine how key ecological metrics vary across depth strata to understand vertical zonation patterns.

```{r}
#| label: fig-transects-summary
#| fig-cap: "Comparison of species richness, total count, density, and biomass across depth strata"

station_summary |> 
  select(ps_station_id, depth_strata, avg_taxa, avg_count, avg_density_m2, avg_biomass_gm2) |>
  pivot_longer(cols = c("avg_taxa", "avg_count", "avg_density_m2", "avg_biomass_gm2"),
               names_to = "metric",
               values_to = "value") |> 
  mutate(metric = recode(metric, 
                         "avg_taxa"        = "Species Richness",
                         "avg_count"       = "Count",
                         "avg_density_m2"  = "Density (ind/m²)",
                         "avg_biomass_gm2" = "Biomass (g/m²)")) |> 
  ggplot(aes(x = depth_strata, y = value, fill = depth_strata)) +
  geom_boxplot(alpha = 0.8, color = "gray40", outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.5, color = "gray30") +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "", 
       y = "",
       fill = "",
       title = "Transect Metrics by Depth Strata",
       subtitle = exp_id) +
  scale_fill_manual(values = ps_colors("depth_strata"))

ggsave(filename = file.path(figures_out, paste0(exp_id, "_fish_metrics_by_subregion.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

### prevalent taxa 

```{r}
#| label: fig-top-taxa
#| fig-cap: "Biomass distributions for top taxa and depth strata"
#| warning: false
#| fig-width: 12
#| fig-height: 8

# Fill missing combinations so absences are treated as zeros
plot_df <- biomass_by_station_and_taxa |> 
  select(ps_station_id, accepted_name, avg_biomass_gm2) |> 
  complete(ps_station_id, 
           accepted_name, 
           fill = list(avg_biomass_gm2 = 0)) |> 
  mutate(accepted_name = forcats::fct_reorder(accepted_name, avg_biomass_gm2, .fun = median, .desc = TRUE)) 

# Identify major taxa by total biomass
taxa_means <- plot_df |> 
  group_by(accepted_name) |> 
  summarise(mean_biomass   = mean(avg_biomass_gm2, na.rm = TRUE),
            median_biomass = median(avg_biomass_gm2, na.rm = TRUE)) |> 
  arrange(desc(median_biomass), desc(mean_biomass)) 

top_taxa <- taxa_means |> 
  slice_head(n = 12) 

# Summarize by depth for medians and means
depth_summary <- plot_df |> 
  left_join(stations |> distinct(ps_station_id, depth_strata), by = "ps_station_id") |> 
  filter(accepted_name %in% top_taxa$accepted_name) |> 
  group_by(accepted_name, depth_strata) |> 
  summarise(mean_biomass_gm2   = mean(avg_biomass_gm2, na.rm = TRUE),
            median_biomass_gm2 = median(avg_biomass_gm2, na.rm = TRUE),
            n_obs = n(),
            .groups = "drop") 

# Plot distributions by depth and taxon
plot_df |> 
  left_join(stations |> select(ps_station_id, depth_strata), by = "ps_station_id") |> 
  filter(accepted_name %in% top_taxa$accepted_name) |> 
  ggplot(aes(x = depth_strata, y = avg_biomass_gm2)) +
  geom_boxplot(aes(color = accepted_name),
               outlier.shape = NA, fill = "white",
               alpha = 0.3, linewidth = 0.6, show.legend = FALSE) +
  geom_jitter(aes(color = accepted_name),
              width = 0.25, size = 1.5, alpha = 0.5, show.legend = FALSE) +
  geom_point(data = depth_summary,
             aes(y = mean_biomass_gm2),
             shape = 23, fill = "black", color = "white",
             size = 3, stroke = 0.8) +
  facet_wrap(~ accepted_name, scales = "free_y", ncol = 4) +
  scale_y_log10(breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
               labels = c("0.001", "0.01", "0.1", "1", "10", "100"),
               expand = expansion(mult = c(0.05, 0.15))) +
  paletteer::scale_color_paletteer_d("ggsci::default_igv") +
  theme(strip.text = element_text(size = 10, face = "italic", lineheight = 1.2, margin = margin(b = 8))) +
  labs(x = NULL,
       y = "Biomass (g/m², log scale)",
       title = "Do dominant taxa show distinct depth preferences?",
       subtitle = exp_id,
       caption = "Top 12 taxa by median biomass • Diamonds = mean • Taxa ordered by overall biomass")

ggsave(filename = file.path(figures_out, paste0(exp_id, "_fish_important_taxa.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

### biomass vs. density

Examine the relationship between biomass and density across trophic groups to understand energy flow.

```{r}
#| label: fig-points-by-taxa
#| fig-cap: "Biomass vs. Density Relationship by Trophic Group"
#| fig-width: 7
#| fig-height: 6

biomass_by_station_and_taxa |> 
  filter(total_count > 0) |> 
  ggplot(aes(x = avg_density_m2, y = avg_biomass_gm2, color = trophic_group)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_density_2d(color = "black", linewidth = 0.2) + # Density contours for highlighting structure
  scale_y_log10(labels = scales::comma_format(accuracy = 0.01)) +
  scale_x_sqrt(labels = scales::comma_format(accuracy = 0.01)) +
  scale_color_manual(values = ps_colors("trophic_group")) +
  labs(x = "Density (ind/m², sqrt scale)", 
       y = "Biomass (g/m², log scale)", 
       color = "",
       title = "Biomass vs. Density Relationship",
       subtitle = exp_id, 
       caption = "Log-transformed biomass vs. square-root-transformed density across trophic groups") 


ggsave(filename = file.path(figures_out, paste0(exp_id, "_fish_biomass_vs_abundance.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

---

# 9. Export & Upload

## Pre-upload summary

```{r}
#| label: data-validation
#| echo: true

# Validate data completeness and basic consistency
validation_summary <- list(
  stations = list(
    total_records      = nrow(station_summary),
    complete_records   = sum(complete.cases(station_summary |> select(-locality, -notes))),
    missing_critical   = sum(is.na(station_summary$ps_station_id) |
                             is.na(station_summary$exp_id) |
                             is.na(station_summary$region)),
    n_regions          = n_distinct(station_summary$region),
    n_subregions       = n_distinct(station_summary$subregion),
    n_depth_strata     = n_distinct(station_summary$depth_strata)
  ),
  
  observations = list(
    total_records       = nrow(biomass_per_obs),
    complete_records    = sum(complete.cases(biomass_per_obs)),
    species_count       = n_distinct(biomass_per_obs$accepted_name),
    largest_fish_cm     = max(biomass_per_obs$length_cm, na.rm = TRUE),
    largest_fish_taxa   = biomass_per_obs$accepted_name[which.max(biomass_per_obs$length_cm)],
    smallest_fish_cm    = min(biomass_per_obs$length_cm, na.rm = TRUE),
    smallest_fish_taxa  = biomass_per_obs$accepted_name[which.min(biomass_per_obs$length_cm)],
    negative_lengths    = sum(biomass_per_obs$length_cm < 0, na.rm = TRUE)
  ),
  
  biomass_summaries = list(
    max_biomass_gm2        = max(station_summary$avg_biomass_gm2, na.rm = TRUE),
    max_biomass_station    = station_summary$ps_station_id[which.max(station_summary$avg_biomass_gm2)],
    max_biomass_subregion  = station_summary$subregion[which.max(station_summary$avg_biomass_gm2)],
    min_biomass_gm2        = min(station_summary$avg_biomass_gm2, na.rm = TRUE),
    min_biomass_station    = station_summary$ps_station_id[which.min(station_summary$avg_biomass_gm2)],
    min_biomass_subregion  = station_summary$subregion[which.min(station_summary$avg_biomass_gm2)],
    zero_biomass_stations  = sum(station_summary$avg_biomass_gm2 == 0, na.rm = TRUE)
  )
)

# Display validation summary
cat("\n📊 DATA VALIDATION SUMMARY\n")
cat("==========================\n\n")

cat("Station Summary:\n")
cat("  • Total records:          ", validation_summary$stations$total_records, "\n")
cat("  • Complete records:       ", validation_summary$stations$complete_records, "\n")
cat("  • Missing critical fields:", validation_summary$stations$missing_critical, "\n")
cat("  • Regions:                ", validation_summary$stations$n_regions, 
    " | Subregions:", validation_summary$stations$n_subregions,
    " | Depth strata:", validation_summary$stations$n_depth_strata, "\n\n")

cat("Observations:\n")
cat("  • Total records:          ", validation_summary$observations$total_records, "\n")
cat("  • Complete records:       ", validation_summary$observations$complete_records, "\n")
cat("  • Unique species:         ", validation_summary$observations$species_count, "\n")
cat("  • Largest fish:           ", validation_summary$observations$largest_fish_cm, " cm (",
    validation_summary$observations$largest_fish_taxa, ")\n")
cat("  • Smallest fish:          ", validation_summary$observations$smallest_fish_cm, " cm (",
    validation_summary$observations$smallest_fish_taxa, ")\n")

cat("Biomass Summaries:\n")
cat("  • Max biomass:            ", validation_summary$biomass_summaries$max_biomass_gm2, " g/m² (",
    validation_summary$biomass_summaries$max_biomass_station, ", ",
    validation_summary$biomass_summaries$max_biomass_subregion, ")\n")
cat("  • Min biomass:            ", validation_summary$biomass_summaries$min_biomass_gm2, " g/m² (",
    validation_summary$biomass_summaries$min_biomass_station, ", ",
    validation_summary$biomass_summaries$min_biomass_subregion, ")\n")
cat("  • Stations with zero biomass:", validation_summary$biomass_summaries$zero_biomass_stations, "\n\n")

cat("✅ Basic validation complete.\n")
```

## Schema Validation

```{r}
validate_and_export <- function(df, table_ref, label, filename) {
  
  ref_cols <- tbl(bq_connection, table_ref) |> 
    filter(!is.na(exp_id)) |> 
    head() |> 
    collect() |> 
    names()
  
  missing <- setdiff(ref_cols, names(df))
  extra <- setdiff(names(df), ref_cols)
  
  if (length(missing) > 0 || length(extra) > 0) {
    cat("\n❌ Schema mismatch for", label, "\n")
    if (length(missing)) cat("  Missing columns:", paste(missing, collapse = ", "), "\n")
    if (length(extra)) cat("  Extra columns:", paste(extra, collapse = ", "), "\n")
    stop("Fix schema before export", call. = FALSE)
  }
  
  out_path <- file.path(data_out, filename)
  df |> select(all_of(ref_cols)) |> write_csv(out_path)
  cat("✓", label, "→", out_path, "\n")
  invisible(out_path)
}

```

## CSV Exports

```{r}
stations_csv <- validate_and_export(df        = station_summary,
                                    table_ref = "uvs.blt_stations",
                                    label     = "Stations",
                                    filename  = paste0(exp_id, "_uvs_blt_stations.csv"))

observations_csv <- validate_and_export(df        = biomass_per_obs |> mutate(notes = NA_character_) ,
                                        table_ref = "uvs.blt_observations",
                                        label     = "Observations",
                                        filename  = paste0(exp_id, "_uvs_blt_observations.csv"))

biomass_csv <- validate_and_export(df        = biomass_by_station_and_taxa,
                                   table_ref = "uvs.blt_biomass_by_station_taxa",
                                   label     = "Biomass by Station | Taxa",
                                   filename  = paste0(exp_id, "_uvs_blt_biomass_by_station_taxa.csv"))
```

## Upload to BigQuery

```{r upload_bq, eval=FALSE}
uploads <- list(
  list(file = stations_csv,     dataset = "uvs", table = "blt_stations"),
  list(file = observations_csv, dataset = "uvs", table = "blt_observations"),
  list(file = biomass_csv,      dataset = "uvs", table = "blt_biomass_by_station_taxa")
)

for (upload in uploads) {
  table_id <- paste0(upload$dataset, ".", upload$table)
  cat("  Uploading:", table_id, "...")
  
  bq_table_upload(
    bq_table("pristine-seas", upload$dataset, upload$table),
    values = read_csv(upload$file, show_col_types = FALSE),
    create_disposition = "CREATE_NEVER",
    write_disposition  = "WRITE_APPEND"
  )
  
  cat("✓\n")
}
```

---