---
title: "Benthos LPI - Data Pipeline"
date: today
format: 
  html:
    theme: minty
    self-contained: true
    code-fold: true
    toc: true 
    toc-depth: 3
    toc-location: right
    html-table-processing: none
execute:
  fig-width: 10
---

```{r setup, message = F, warning = F, fig.width = 10, fig.height = 10, echo = F}
options(scipen = 999)

# Hook to format inline numeric expressions with comma separators:
knitr::knit_hooks$set(inline = function(x) {
  # For non-numeric values, just return as character
  if (!is.numeric(x)) {return(as.character(x))}
  # Format numbers with comma as big.mark, no scientific notation
  else format(x, big.mark = ",", scientific = FALSE)})

library(sf)
library(hms)
library(readxl)
library(janitor)
library(lubridate)
library(gt)
library(pointblank)
library(tidyverse)
library(bigrquery)
library(leaflet)
library(leaflet.extras)
library(ggtext)
library(PristineSeasR2)

# ---- Parameters -----------------------------------------------------------

exp_id   <- "ANT_2019"

ps_paths <- PristineSeasR2::get_drive_paths()

prj_path <- file.path(ps_paths$projects, "legacy-db")

# ---- Auth & Connection ----------------------------------------------------

bigrquery::bq_auth(email = "marine.data.science@ngs.org")
bq_connection <- DBI::dbConnect(bigrquery::bigquery(), project = "pristine-seas")

set_theme(PristineSeasR2::theme_pristineseas())

data_out <- file.path(prj_path, "data/output/uvs")

figures_out <- file.path(prj_path, "figures/uvs")

legacy_path <- file.path(ps_paths$science, "legacy data/Antarctica/")
```

# Overview

This script outlines the end-to-end pipeline for processing Line Point Intercept (LPI) survey data collected during Pristine Seas expeditions. The pipeline ingests raw field data, standardizes formats, performs taxonomy lookups, applies quality assurance/quality control (QA/QC), computes station-level summaries, explores pattenrs, and loads the clean data into the Pristine Seas Science Database in BigQuery.

# 1. Data Ingestion

## Reference Tables

We begin by reading the UVS (Underwater Visual Survey) site information already processed.
  
```{r load_references}
uvs_sites <- tbl(bq_connection, "uvs.sites") |> 
   filter(exp_id == !!exp_id) |> 
   collect()

# Validate site data
cat("Sites loaded:", nrow(uvs_sites), "\n")
cat("Regions:",      unique(uvs_sites$region), "\n")
cat("Subregions:",   unique(uvs_sites$subregion), "\n")

# Benthic taxonomy reference
benthos_taxa_lut <- tbl(bq_connection, "taxa_info.benthos") |> 
  filter(!is.na(accepted_name)) |> 
  collect()

cat("✓ Loaded", nrow(benthos_taxa_lut), "benthic taxa\n")
```

## Fieldbooks

```{r}
transects_raw <- read_xlsx(file.path(legacy_path, "Benthos/New_Kike/Benthos Antarctica.xlsx"),
                          range         = "A1:L3",
                          col_names     = FALSE,
                          .name_repair  = "minimal") |>
  t() |>
  as_tibble(.name_repair = "minimal") |> 
  row_to_names(row_number = 1) |>
  clean_names() |> 
  mutate(depth_m          = parse_number(depth),
         site             = str_pad(parse_number(station), 3, pad = "0"),
         ps_site_id       = paste(exp_id, "uvs", site, sep = "_")) |> 
  select(ps_site_id, depth_m)

transects <- transects_raw |> 
  mutate(exp_id         = exp_id,
         survey_type    = "uvs",
         method         = "lpi",
         diver          = "Kike Ballesteros",
         transect       = "A",
         transect_length_m = 25,
         depth_strata   = stratify(depth_m),
         ps_station_id  = paste(ps_site_id, station_suffix(depth_strata), sep = "_")) |> 
  select(exp_id, survey_type, ps_site_id, ps_station_id, method, diver, 
         depth_m, depth_strata, transect, transect_length_m)

contacts_raw <- read_xlsx(file.path(legacy_path, "Benthos/New_Kike/Benthos Antarctica.xlsx"),
                          range         = "A6:L45",
                          col_names     = FALSE,
                          .name_repair  = "minimal") 

contacts <- contacts_raw |>
    set_names(c("morphotaxon", transects$ps_station_id)) |>
    mutate(across(all_of(transects$ps_station_id), as.character)) |>
    pivot_longer(cols      = -c(morphotaxon),  # Exclude BOTH taxonomy columns
                 names_to  = "ps_station_id",
                 values_to = "contacts") |>
    mutate(contacts = parse_number(contacts)) |>
    filter(!is.na(contacts)) |>
    select(ps_station_id, morphotaxon, contacts)

transect_totals <- contacts |>
    group_by(ps_station_id) |>
    summarise(n_points = sum(contacts, na.rm = TRUE),
              .groups  = "drop")
  
lpi_transects <- transects |>
  left_join(transect_totals, by = "ps_station_id")

lpi_contacts <- contacts |>
  left_join(transects |> 
              select(exp_id, ps_station_id, diver, depth_m, transect, transect_length_m), 
            by = "ps_station_id") |>
   select(ps_station_id, exp_id, diver, depth_m, transect, transect_length_m, morphotaxon, count = contacts) |> 
  arrange(ps_station_id, transect) 
```

# 3. Taxonomy Resolution

```{r}
# Summarize by mophotaxon

lpi_taxa <- lpi_contacts |> 
  group_by(morphotaxon) |> 
  summarise(count = round(sum(count), 2),
            .groups = "drop") |> 
  mutate(pct_count = round(100 * count / sum(count), 2)) |> 
  filter(!morphotaxon %in% c("Barren", "Sediment")) |> 
  arrange(desc(count))

lpi_taxa <- lpi_taxa |> 
  mutate(taxon_clean = bdc::bdc_clean_names(morphotaxon)$names_clean) 

lpi_taxa$taxon_clean[lpi_taxa$morphotaxon == "CCA"] <- "Corallinales"
lpi_taxa$taxon_clean[lpi_taxa$morphotaxon == "Terebellidae unidentified"] <- "Terebellidae"
lpi_taxa$taxon_clean[lpi_taxa$morphotaxon == "Unidentified bryozoan flabellate"] <- "Bryozoa"
lpi_taxa$taxon_clean[lpi_taxa$morphotaxon == "Hormosira scotti"] <- "Hormosoma scotti"

lpi_taxa |> filter(is.na(taxon_clean))
```

```{r}
# Let's check against our Taxonomic Reference Table

# Create name-to-AphiaID mapping

names_to_aphiaID <- benthos_taxa_lut |> 
  select(accepted_aphia_id, taxon_name, accepted_name) |>
  pivot_longer(cols = c("taxon_name", "accepted_name"), names_to = "orig_col", values_to = "lookup_name") |>
  filter(!is.na(lookup_name)) |>
  distinct(accepted_aphia_id, lookup_name) 

lpi_taxa <- lpi_taxa |> 
  left_join(names_to_aphiaID, 
            by = c("taxon_clean" = "lookup_name")) 
  
lpi_taxa |> 
  filter(is.na(accepted_aphia_id))
```

```{r eval = F}
new_taxa <- lpi_taxa |> 
  filter(is.na(accepted_aphia_id)) |> 
  select(morphotaxon, taxon_clean, pct_count)

new_worms_records <- purrr::map_dfr(new_taxa$taxon_clean, 
                            ~worrms::wm_records_names(.x)) |> 
  select(taxon_clean = scientificname, aphia_id = AphiaID, rank, 
         name_status = status, accepted_name = valid_name, accepted_aphia_id = valid_AphiaID) |> 
  mutate(rank = str_to_lower(rank))

dupes <- new_worms_records |> 
  get_dupes(taxon_clean)

deduped <- dupes |> 
  filter(name_status == "accepted") |> 
  select(-dupe_count) |> 
  distinct()

new_worms_records <- new_worms_records |>
  filter(!taxon_clean %in% dupes$taxon_clean) |> 
  bind_rows(deduped)

get_taxonomic_ranks <- function(id) {
  tryCatch(
    {worrms::wm_classification(id) |>
        select(rank, scientificname) |>
        pivot_wider(names_from = rank, values_from = scientificname) |>
        mutate(accepted_aphia_id = id)},
    error = function(e) {
      tibble(accepted_aphia_id = id)
    })
  }

library(furrr)

new_taxonomy_ranks <- furrr::future_map_dfr(new_worms_records$accepted_aphia_id,
                                            get_taxonomic_ranks,
                                            .options = furrr_options(seed = TRUE)) |>
  clean_names() |>
  select(accepted_aphia_id, kingdom, phylum, class, order, family, genus)

new_taxa <- new_taxa |> 
  left_join(new_worms_records) |> 
  left_join(new_taxonomy_ranks) |> 
  rename(taxon_name = taxon_clean,
         status = name_status) |> 
  select(-morphotaxon, -pct_count) 

new_taxa$phylum[new_taxa$class == "Florideophyceae" & is.na(new_taxa$phylum)] <- "Chlorophyta"
new_taxa$phylum[new_taxa$class == "Ulvophyceae" & is.na(new_taxa$phylum)] <- "Chlorophyta"

# Antarctic benthic taxa mapping to standard functional groups ----

new_taxa <- new_taxa |>
  mutate(functional_group = case_when(
    genus %in% c("Desmarestia", "Himantothallus", "Ascoseira") ~ "kelp",
    genus %in% c("Cystosphaera", "Halopteris", "Adenocystis") ~ "algae_erect",
    genus %in% c("Iridaea", "Phyllophora") ~ "algae_encrusting",
    genus %in% c("Gigartina", "Ballia", "Trematocarpus", "Neuroglossum",
                 "Myriogramme", "Pantoneura", "Georgiella", "Phycodrys") ~ "algae_erect",
    genus %in% c("Cinachyra", "Polymastia", "Sphaerotylus",
                 "Dendrilla", "Isodictya", "Mycale") ~ "sponges",
    genus %in% c("Cnemidocarpa", "Molgula", "Synoicum") ~ "ascidians",
    genus %in% c("Onogorgia") ~ "soft_coral",
    genus %in% c("Hormosoma") ~ "other_cnidarians",
    genus %in% c("Nematoflustra") ~ "bryozoans",
    genus %in% c("Paralia") ~ "algae_encrusting",
    family %in% c("Terebellidae") ~ "worms",
    genus %in% c("Heterocucumis") ~ "other",
    TRUE ~ NA
  ))


# IUCN status

iucn_db <- read_csv(file.path(ps_paths$datasets, 
                              "iucn-redlist-marine-species",  
                              "joined_and_resolved_taxa.csv"))

iucn_priority <- c("CR", "EN", "VU", "NT", "LC", "DD", "NE")

iucn_clean <- iucn_db |> 
  rename(accepted_name = taxon_valid_name) |> 
  filter(accepted_name %in% new_taxa$accepted_name) |> 
  distinct(accepted_name, iucn_redlist_cat, iucn_taxon_id) |> 
  mutate(iucn_redlist_cat = case_when(str_detect(iucn_redlist_cat, "Critically Endangered") ~ "CR",
                                      str_detect(iucn_redlist_cat, "Endangered") ~ "EN",
                                      str_detect(iucn_redlist_cat, "Vulnerable") ~ "VU",
                                      str_detect(iucn_redlist_cat, "Near Threatened") ~ "NT",
                                      str_detect(iucn_redlist_cat, "Least Concern") ~ "LC",
                                      str_detect(iucn_redlist_cat, "Data Deficient") ~ "DD",
                                      str_detect(iucn_redlist_cat, "Extinct") ~ "EX",
                                      str_detect(iucn_redlist_cat, "Lower Risk/least concern") ~ "LC",
                                      str_detect(iucn_redlist_cat, "Lower Risk/near threatened") ~ "NT",
                                      TRUE ~ NA_character_)) |> 
  mutate(iucn_rank = match(iucn_redlist_cat, c("EX", iucn_priority))) |> 
  group_by(accepted_name) |> 
  arrange(iucn_rank) |> 
  slice(1) |> 
  ungroup() |> 
  select(-iucn_rank) |> 
  rename(iucn_cat = iucn_redlist_cat,
         iucn_sis_id = iucn_taxon_id)

new_taxa <- new_taxa |> 
  left_join(iucn_clean)

# upload to BQ

bq_table_upload(bq_table("pristine-seas", "taxa_info", "benthos"),
                values = new_taxa,
                create_disposition = "CREATE_NEVER",
                write_disposition = "WRITE_APPEND")
```

```{r}
lpi_taxa <- lpi_taxa |> 
  left_join(benthos_taxa_lut |> 
              filter(!aphia_id %in% c(224174, 1341, 224179), !is.na(accepted_aphia_id)) |> 
              distinct(accepted_aphia_id, accepted_name, rank, functional_group, family)) 
```

```{r}
lpi_taxa |> 
  mutate(rank = coalesce(rank, morphotaxon)) |> 
  group_by(rank) |> 
  summarize(n_morphotaxa = n_distinct(morphotaxon),
            n_aphia_id = n_distinct(accepted_aphia_id),
            contacts = sum(count),
            .groups = "drop") |> 
  arrange(desc(contacts)) |> 
  mutate(pct_contacts = round(100*contacts/sum(contacts), 2))

lpi_taxa |> 
  filter(!is.na(family)) |> 
  group_by(family) |> 
  summarize(n_morphotaxa = n_distinct(morphotaxon),
            n_aphia_id = n_distinct(accepted_aphia_id),
            contacts = sum(count),
            .groups = "drop") |> 
  arrange(desc(contacts)) |> 
    mutate(pct_contacts = round(100*contacts/sum(contacts), 2))

lpi_taxa |> 
  group_by(functional_group) |> 
  summarize(n_morphotaxq = n_distinct(morphotaxon),
            n_aphia_id = n_distinct(accepted_aphia_id),
            contacts = sum(count),
            .groups = "drop") |> 
  arrange(desc(contacts)) |> 
  mutate(pct_contacts = round(100*contacts/sum(contacts), 2)) # Excluding the non taxa groups e.g., barren

lpi_taxa |> filter(is.na(functional_group))
```

## Join Taxonomy to Observations

```{r}
# Combine with contacts data

lpi_observations <- lpi_contacts |> 
  left_join(lpi_taxa |> 
              distinct(morphotaxon, accepted_name, accepted_aphia_id, rank, functional_group, family), 
            by = "morphotaxon") |> 
  mutate(functional_group = case_when(morphotaxon %in% c("Barren", "Sediment", "Sand") ~ "sediment | rubble | barren", 
                                      morphotaxon == "Turf" ~ "turf",
                                      TRUE ~ functional_group),
         notes = NA_character_) |> 
  select(ps_station_id, exp_id, diver, depth_m, transect, transect_length_m, 
         morphotaxon, accepted_name, accepted_aphia_id, rank, family, functional_group, count, notes) 

lpi_observations |> 
  get_dupes(ps_station_id, transect, morphotaxon)
```

# 4. Summarize  Cover

## By Transect

```{r}
cover_by_transect_taxa <- lpi_observations |>
  mutate(transect_section = "0-25") |> 
  # 1. Section-level % cover (for SD)
  group_by(ps_station_id, transect, transect_section) |>
  mutate(pct_cover_section = 100 * count / sum(count)) |>
  ungroup() |>
  # 2. Fill missing taxa × section combos
  group_by(ps_station_id, transect, diver) |>
  complete(nesting(transect_section),
           nesting(functional_group, morphotaxon, accepted_name, accepted_aphia_id, rank, family),
           fill = list(count             = 0, 
                       pct_cover_section = 0)) |>
  ungroup() |>
  # 3. Compute total contacts and SD per transect (sum across sections)
  group_by(diver, ps_station_id, transect, 
           morphotaxon, accepted_name, accepted_aphia_id, functional_group, rank, family) |>
  summarize(total_count    = sum(count),
            pct_cover_sd   = sd(pct_cover_section),
            .groups = "drop") |>
  # 4. Compute compositional % cover
  group_by(ps_station_id, transect) |>
  mutate(pct_cover = round(100 * total_count / sum(total_count), 4)) |>
  ungroup() 
```

## By Station

```{r}
lpi_stations <- lpi_transects |> 
  group_by(ps_station_id, depth_strata) |> 
  summarize(divers        = paste0(sort(unique(diver), decreasing = T), collapse = " | "),
            depth_m       = round(mean(depth_m)),
            n_transects   = n_distinct(transect),
            n_sections    = 1,
            n_points      = sum(n_points),
            survey_dist_m = sum(transect_length_m),
            .groups = "drop") |> 
  # Add # Taxa
  left_join(lpi_observations |>
              filter(! morphotaxon %in% c("Barren", "Sediment", "Turf")) |> 
              group_by(ps_station_id) |>
              summarize(n_taxa = n_distinct(accepted_name[count > 0 ])),
            by = "ps_station_id") |> 
  # Add site info
  mutate(ps_site_id = str_replace(ps_station_id, "_[^_]+$", "")) |> 
  left_join(uvs_sites |> 
              distinct(exp_id, ps_site_id, region, subregion, habitat, locality, exposure),
            by = "ps_site_id") |> 
  select(ps_station_id, ps_site_id, exp_id, region, subregion, locality, habitat, exposure, 
         depth_strata, depth_m,  divers, n_transects, survey_dist_m, n_points, n_taxa)
```

```{r}
cover_by_station_taxa <- cover_by_transect_taxa |> 
  select(ps_station_id, transect, morphotaxon, accepted_name, functional_group, total_count, pct_cover, pct_cover_sd) |> 
  group_by(ps_station_id) |> 
  complete(nesting(transect), 
           nesting(morphotaxon, accepted_name, functional_group),
           fill = list(total_count  = 0,
                       pct_cover    = 0,
                       pct_cover_sd = NA_real_)) |> 
  group_by(ps_station_id, morphotaxon, accepted_name, functional_group,) |> 
  summarize(total_count  = round(sum(total_count), 3),
            pct_cover    = round(mean(pct_cover),3),
            pct_cover_sd = round(mean(pct_cover_sd),3),
            .groups = "drop") 

cover_by_station_taxa <- lpi_stations |> 
  inner_join(cover_by_station_taxa) |> 
  left_join(lpi_taxa |> 
              distinct(morphotaxon, accepted_name, accepted_aphia_id, rank, family, functional_group)) |> 
  select(ps_station_id, ps_site_id, exp_id, region, subregion, habitat, exposure, 
         depth_strata, depth_m, divers, n_transects, survey_dist_m,
         morphotaxon, accepted_name, accepted_aphia_id, rank, family, functional_group, total_count, pct_cover, pct_cover_sd)
```

```{r}
others_group <- c("forams","bryozoans","worms","echinoderms", "molluscs", "hydrozoans", "ascidians", "other_cnidarians", "worms")

grps_order <- c("cyanobacteria", "eam", "sediment | rubble | barren", "other", "turf", "algae_encrusting", "algae_erect","sponges", "soft_coral","cca", "hard_coral")

cover_by_station_group <- cover_by_station_taxa |> 
  mutate(functional_group = case_when(functional_group %in% others_group         ~ "other",
                                      functional_group %in% c("hard_coral_dead") ~ "sediment | rubble | barren",
                                      TRUE ~ functional_group)) |> 
  group_by(region, subregion, ps_site_id, ps_station_id, depth_strata, functional_group) |> 
  summarize(pct_cover = round(sum(pct_cover),2),
            .groups   = "drop") |> 
  pivot_wider(names_from = functional_group, 
              values_from = pct_cover, 
              values_fill = 0) |> 
  pivot_longer(cols = -c(ps_station_id, region, subregion, depth_strata, ps_site_id), 
               names_to = "functional_group", 
               values_to = "pct_cover") |>
  mutate(functional_group = fct_relevel(functional_group, grps_order))

cover_by_station_group_wide <- cover_by_station_group |> 
  select(ps_station_id, functional_group, pct_cover) |>
  pivot_wider(names_from = functional_group, 
              values_from = pct_cover, 
              values_fill = 0,
              names_prefix = "pct_") |> 
  janitor::clean_names() |> 
  rename(pct_algae_encrust = pct_algae_encrusting,
         pct_rubble        = pct_sediment_rubble_barren)

lpi_stations <- lpi_stations |> 
  left_join(cover_by_station_group_wide)
```

# 5. QA/QC Validation

## Transect Checks

```{r}
agent_transects <- lpi_transects |> 
  create_agent(label = "LPI Transects QA/QC",
               tbl_name = "transects") |> 
  # Core fields must be complete
  rows_complete(columns = vars(ps_station_id, transect, diver, transect_length_m, n_points),
                label = "No missing core fields",
                actions = action_levels(stop_at = 0.01)) |> 
  # Survey effort was standard
  col_vals_equal(columns = vars(transect_length_m),
                   value = 50,
                   label = "Five sections per transect",
                   actions = action_levels(warn_at = 0.1)) |>
  col_vals_equal(columns = vars(n_points),
                   value = 125,
                   label = "250 points per transect",
                   actions = action_levels(warn_at = 0.1)) |>
  interrogate()

agent_transects
```

## Station Checks

```{r}
# Station QA/QC using pointblank
station_qaqc <- lpi_stations |> 
  create_agent(label = "Benthos LPI Stations QA/QC", tbl_name = "lpi_stations") |> 
  rows_distinct(ps_station_id,
                label = "Station IDs are unique",
                actions = action_levels(stop_at = 0.001)) |>
  col_vals_equal(columns = vars(n_transects), 
                 value = 1,
                 label = "Expected 1 transects per station",
                 actions = action_levels(warn_at = 0.001)) |>
  interrogate()

# Display QA/QC results
station_qaqc
```

```{r}
cover_by_station_taxa |> 
  group_by(ps_station_id) |> 
  summarize(total_cover = round(sum(pct_cover))) |> 
  filter(total_cover != 100)
```

## Observation Checks

```{r}
cover_by_station_taxa |> 
  filter(is.na(accepted_name), 
         !functional_group %in% c("sediment | rubble | barren", "turf"))
```

```{r}
lpi_observations |> 
  create_agent(label = "Benthos LPI Observations QA/QC", tbl_name = "lpi_observations") |> 
  # funcitonal groups in allowed vocab
  col_vals_in_set(columns = vars(functional_group),
                  set     = allowed_vocab$functional_groups,
                  label   = "Functional group in allowed vocabulary",
                  actions = action_levels(stop_at = 0.01)) |>
  interrogate()
```


# 6. Summary Statistics

```{r summary_stats}
#| label: tbl-summary
#| tbl-cap: "LPI survey effort by region and depth strata"

lpi_stations |> 
  group_by(region, depth_strata) |> 
  summarise(n_stations      = n(),
            mean_taxa       = round(mean(n_taxa), 1),
            mean_pct_kelp    = round(mean(pct_kelp), 1),
            .groups = "drop") |> 
  pivot_wider(names_from  = depth_strata,
              values_from = c(n_stations, mean_taxa, mean_pct_kelp),
              values_fill = 0)  |> 
  gt(groupname_col = "region") |> 
  tab_spanner(label = "Stations", columns = starts_with("n_stations")) |> 
  tab_spanner(label = "Mean Taxa", columns = starts_with("mean_taxa")) |> 
  tab_spanner(label = "Mean % kelp", columns = starts_with("mean_pct_kelp")) |> 
  cols_label(region = "region",
             n_stations_deep            = "Deep",
             n_stations_shallow         = "Shallow",
             mean_taxa_deep             = "Deep",
             mean_taxa_shallow          = "Shallow",
             mean_pct_kelp_deep          = "Deep",
             mean_pct_kelp_shallow       = "Shallow") |> 
  tab_options(table.font.size = 12, row_group.as_column = TRUE)
```

```{r}
lpi_stations |> 
  filter(exposure != "lagoon") |> 
  group_by(region) |> 
  summarise(n_stations         = n_distinct(ps_station_id),
            avg_n_taxa         = mean(n_taxa),
            across(contains("pct"), mean), 
            .groups = "drop") |> 
  mutate_if(is.numeric, round, 1)
```

```{r}
tmp <- cover_by_station_taxa |> 
  filter(functional_group != "sediment | rubble | barren") |> 
  mutate(accepted_name = coalesce(accepted_name, morphotaxon)) |>
  select(ps_station_id, accepted_name, total_count, pct_cover) |> 
  complete(nesting(ps_station_id),
           nesting(accepted_name),
           fill = list(pct_cover = 0,
                       total_count = 0))
```

```{r taxa_summary}
#| label: tbl-taxa
#| tbl-cap: "Top 10 invertebrate taxa by frequency"

taxa_summary <- cover_by_station_taxa |> 
  filter(functional_group != "sediment | rubble | barren") |> 
  mutate(accepted_name = coalesce(accepted_name, morphotaxon)) |>
  select(ps_station_id, accepted_name, total_count, pct_cover) |> 
  complete(nesting(ps_station_id),
           nesting(accepted_name),
           fill = list(pct_cover = 0,
                       total_count = 0)) |> 
  group_by(accepted_name) |> 
  summarize(total_count = round(sum(total_count)),
            freq_obs    = round(n_distinct(ps_station_id[pct_cover > 0])/length(unique(lpi_stations$ps_station_id)),3),
            mean_cover  = round(mean(pct_cover), 2),
            .groups = "drop") |> 
  arrange(desc(mean_cover))

taxa_summary |> 
  slice_head(n = 10) |> 
  gt() |> 
  cols_label(accepted_name = "Scientific Name",
             total_count   = "Total Count",
             freq_obs      = "Fraction Stations Observed",
             mean_cover  = "Mean Cover (%)") |> 
  fmt_number(columns = c(mean_cover), decimals = 2) |> 
  tab_options(table.font.size = 12)
```

# 7. Map

```{r}
#| label: fig-map
#| fig-cap: "Interactive map of LPI survey stations"

# Create spatial features
lpi_sites_sf <- uvs_sites |> 
  select(ps_site_id, region, subregion, locality, habitat, exposure, latitude, longitude) |> 
  inner_join(lpi_stations |> 
               group_by(ps_site_id) |> 
               summarize(divers = paste(unique(divers), collapse = "/"),
                         strata = paste(unique(paste(depth_strata, " (", depth_m, "m)", sep = "")), 
                                        collapse = "\n"),
                         survey_dist = sum(survey_dist_m),
                         n_stations = n_distinct(ps_station_id),
                         n_taxa = mean(n_taxa),
                         pct_kelp = mean(pct_kelp),
                         .groups = "drop"),
             by = "ps_site_id") |> 
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 

# Create interactive map
mapview::mapview(lpi_sites_sf,
                 zcol = "pct_kelp",
                 legend = TRUE,
                 map.types = "Esri.WorldImagery",
                 layer.name = "pct_kelp",
                 popup = leafpop::popupTable(lpi_sites_sf, 
                                             zcol = c("ps_site_id", "strata", "divers", "survey_dist", "habitat", "exposure"))) |> 
  leafem::addMouseCoordinates() |> 
  addFullscreenControl()
```

# 8. Exploratory Visualizations

## Site composition

```{r}
cover_by_station_group |> 
  group_by(region, ps_site_id, depth_strata, functional_group) |> 
  summarize(pct_cover = round(sum(pct_cover),2),
            .groups = "drop") |> 
  mutate(site = str_extract(ps_site_id, "[^_]+$")) |> 
  filter(depth_strata != "supershallow") |> 
  ggplot()+
  geom_col(aes(site, pct_cover, fill = functional_group))+
  facet_grid(depth_strata ~ region, scales = "free")+
  scale_fill_manual(values = c(ps_colors("functional_groups"), "kelp" = "brown"))+
  labs(x = NULL, 
       y = "Benthic composition (% of station total)",
       fill = "Functional group",
       title = paste(exp_id, ":","Benthic Cover by Site"),
       subtitle = "Composition by depth strata")

ggsave(filename = file.path(figures_out, paste0(exp_id, "_lpi_cover_by_site.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

## Subregion composition

```{r}
cover_by_station_group |> 
  group_by(region, subregion, depth_strata, functional_group) |> 
  summarize(pct_cover = mean(pct_cover)) |> 
  ungroup() |> 
  filter(depth_strata != "supershallow") |> 
  ggplot()+
  geom_col(aes(subregion, pct_cover, fill = functional_group))+
  facet_grid(depth_strata~region, scales = "free")+
  scale_fill_manual(values = c(ps_colors("functional_groups"), "kelp" = "brown"))+
  labs(x = NULL, 
       y = "Benthic composition (% of station total)",
       fill = "Functional group",
       title = paste(exp_id, ":", "Benthic Cover by Subregion"),
       subtitle = "Composition by depth strata")

ggsave(filename = file.path(figures_out, paste0(exp_id, "_lpi_cover_by_subregion.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

## Patterns by depth strata

```{r}
#| label: fig-transects-summary
#| fig-cap: "Comparison of benthic cover across depth strata showing how reef composition changes with depth"

lpi_stations |> 
  select(ps_station_id, depth_strata, pct_kelp, pct_cca, pct_algae_erect, pct_rubble) |> 
  pivot_longer(cols = -c(ps_station_id, depth_strata),
               names_to = "functional_group",
               values_to = "pct_cover") |> 
  mutate(functional_group = case_when(functional_group == "pct_kelp" ~ "Kelp",
                                      functional_group == "pct_cca" ~ "Crustose Coralline Algae",
                                      functional_group == "pct_algae_erect" ~ "Erect Macroalgae",
                                      functional_group == "pct_rubble" ~ "Rubble",
                                      TRUE ~ functional_group),
         functional_group = factor(functional_group, 
                                   levels = c("Kelp", "Crustose Coralline Algae",
                                              "Erect Macroalgae",  "Rubble"))) |> 
  ggplot(aes(x = depth_strata, y = pct_cover, fill = depth_strata)) +
  geom_boxplot(alpha = 0.65, 
               color = "gray30", 
               outlier.shape = NA,
               linewidth = 0.5,
               show.legend = FALSE) +
  geom_jitter(width = 0.25, 
              size = 1.2, 
              alpha = 0.35, 
              color = "gray20",
              show.legend = FALSE) +
  facet_wrap(~functional_group, 
             scales = "free_y",
             ncol = 3) +
  labs(x = NULL, 
       y = "Percent Cover (%)",
       title = paste(exp_id, ":", "Reef Benthic Composition Across Depth Strata"),
       subtitle = "Distribution of functional group cover by depth strata • Boxes show median and quartiles • Points represent individual stations") +
  scale_fill_manual(values = ps_colors("depth_strata"))

ggsave(filename = file.path(figures_out, paste0(exp_id, "_lpi_patterns_by_depth.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

## % Hard coral

```{r}
library(ggbeeswarm)

lpi_stations |> 
  ggplot(aes(x = subregion, y = pct_kelp)) +
  geom_quasirandom(aes(color = pct_kelp), 
                   alpha = 0.4, 
                   size = 1.5,
                   width = 0.3,show.legend = F) +
  stat_summary(fun = mean, geom = "point", 
               shape = 23, size = 3, 
               fill = "black", color = "white", stroke = 1) +
  stat_summary(fun = median, geom = "crossbar",
               width = 0.5, linewidth = 0.4, color = "black") +
  facet_wrap(~region, scales = "free_x", ncol = 4) +
  labs(title = paste(exp_id, ":", "Hard Coral Cover by Subregions"),
       caption = "Each dot represents a survey station • Black diamonds show subregion averages • Gray bars indicate median values",
       x = NULL,
       y = "Hard Coral Cover (%)",
       color = "Coral Cover (%)")

ggsave(filename = file.path(figures_out, paste0(exp_id, "_lpi_coral_cover_by_subregion.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

## Most abundant taxa

```{r warning = F}
#| label: fig-top-taxa
#| fig-cap: "Percent cover distributions for the 12 most abundant taxa across depth strata, showing median and mean values with log-transformed scales to accommodate the wide range of cover values"
#| fig-width: 14
#| fig-height: 10

library(ggtext)

# Identify major taxa based on average % cover (after filling with zeros)

plot_df <- cover_by_station_taxa |> 
  select(ps_station_id, morphotaxon, accepted_name, functional_group, 
         accepted_aphia_id, rank, pct_cover, total_count) |> 
  complete(ps_station_id, 
           nesting(morphotaxon, accepted_name, functional_group, accepted_aphia_id, rank),
           fill = list(pct_cover = 0, total_count = 0))

taxa_means <- plot_df |> 
  filter(!is.na(accepted_name)) |> 
  group_by(morphotaxon, accepted_name) |> 
  summarise(mean_cover = mean(pct_cover, na.rm = TRUE),
            median_cover = median(pct_cover, na.rm = TRUE),
            total_cover = sum(pct_cover, na.rm = TRUE),
            .groups = "drop") |> 
  arrange(desc(median_cover), desc(mean_cover)) 

top_taxa <- taxa_means |> 
  head(12) |> 
  pull(morphotaxon)

# Add species names for better labels
top_taxa_names <- taxa_means |> 
  filter(morphotaxon %in% top_taxa) |> 
  mutate(# Create cleaner display names with scientific names in italics
         display_name = paste0("\n<i>", accepted_name, "</i>"),
         # Order by abundance for logical facet arrangement
         display_name = forcats::fct_reorder(display_name, median_cover, .desc = TRUE)) |> 
  select(morphotaxon, display_name)

# Compute summary stats for annotation
depth_summary <- plot_df |> 
  left_join(lpi_stations, by = "ps_station_id") |> 
  filter(morphotaxon %in% top_taxa) |> 
  group_by(morphotaxon, depth_strata) |> 
  summarise(mean_cover = mean(pct_cover, na.rm = TRUE),
            median_cover = median(pct_cover, na.rm = TRUE),
            n_obs = n(),
            .groups = "drop") |> 
  left_join(top_taxa_names, by = "morphotaxon")

# Plot

# Main plot
plot_df |> 
  left_join(lpi_stations, by = "ps_station_id") |> 
  filter(morphotaxon %in% top_taxa) |> 
  left_join(top_taxa_names, by = "morphotaxon") |> 
  ggplot(aes(x = depth_strata, y = pct_cover + 0.001)) +  # Add small constant for log scale
  # Boxplot (cleaner without fill, let color do the work)
  geom_boxplot(aes(color = morphotaxon), 
               outlier.shape = NA, 
               alpha = 0.3,
               fill = "white",
               linewidth = 0.6,
               show.legend = FALSE) +
  # Jittered points
  geom_jitter(aes(color = morphotaxon), 
              width = 0.25, 
              size = 1.5, 
              alpha = 0.5,
              show.legend = FALSE) +
  # Median as bold line segment
  geom_segment(data = depth_summary,
               aes(x = as.numeric(depth_strata) - 0.35,
                   xend = as.numeric(depth_strata) + 0.35,
                   y = median_cover + 0.001,
                   yend = median_cover + 0.001),
               color = "black",
               linewidth = 1.2,
               alpha = 0.8) +
  # Mean as diamond point
  geom_point(data = depth_summary,
             aes(y = mean_cover + 0.001),
             shape = 23,
             fill = "#e74c3c",
             color = "white",
             size = 3,
             stroke = 0.8) +
  # Facet by taxa with scientific names
  facet_wrap(~ display_name, 
             scales = "free_y",
             ncol = 4) +
  # Log scale with cleaner breaks
  scale_y_log10(breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
                labels = c("0", "0.01", "0.1", "1", "10", "100"),
                expand = expansion(mult = c(0.05, 0.15))) +
  # Labels with clear narrative
  labs(x = NULL, 
       y = "Percent Cover (%, log scale)", 
       title    = paste(exp_id, ":", "Dominant Coral Taxa Show Distinct Depth Preferences and Cover Variability"),
       subtitle = "Top 12 taxa by median abundance • Bold lines show median • Diamonds show mean • Taxa ordered by overall abundance",
       caption  = "Note: Y-axis uses log scale to visualize the wide range of cover values (0.01% to 100%)") +
  paletteer::scale_fill_paletteer_d("ggsci::default_igv") +
  paletteer::scale_color_paletteer_d("ggsci::default_igv")+
  theme(strip.text = element_markdown(size = 10, face = "bold",
                                   lineheight = 1.2,
                                   margin = margin(b = 8)))

ggsave(filename = file.path(figures_out, paste0(exp_id, "_lpi_important_taxa.png")),
       width = 14, height = 7, dpi = 300, bg = "white")
```

---

# 9. Export & Upload

## Pre-upload summary

## Schema Validation

```{r validate_schema}
validate_and_export <- function(df, table_ref, label, filename) {
  
  ref_cols <- tbl(bq_connection, table_ref) |> 
    filter(!is.na(exp_id)) |> 
    head() |> 
    collect() |> 
    names()
  
  missing <- setdiff(ref_cols, names(df))
  extra <- setdiff(names(df), ref_cols)
  
  if (length(missing) > 0 || length(extra) > 0) {
    cat("\n❌ Schema mismatch for", label, "\n")
    if (length(missing)) cat("  Missing columns:", paste(missing, collapse = ", "), "\n")
    if (length(extra)) cat("  Extra columns:", paste(extra, collapse = ", "), "\n")
    stop("Fix schema before export", call. = FALSE)
  }
  
  out_path <- file.path(data_out, filename)
  df |> select(all_of(ref_cols)) |> write_csv(out_path)
  cat("✓", label, "→", out_path, "\n")
  invisible(out_path)
}
```

## CSV Exports

```{r}
stations_csv <- validate_and_export(
  df        = lpi_stations |> mutate(pct_coral = NA,
                                     pct_cyano = NA, 
                                     pct_turf = NA, pct_eam = NA, pct_seagrass = NA, notes = NA),
  table_ref = "uvs.lpi_stations",
  label     = "Stations",
  filename  = paste0(exp_id, "_uvs_lpi_stations.csv")
)

observations_csv <- validate_and_export(
  df        = lpi_observations |> mutate(transect_section = "0-25"),
  table_ref = "uvs.lpi_observations",
  label     = "Observations",
  filename  = paste0(exp_id, "_uvs_lpi_observations.csv")
)

cover_csv <- validate_and_export(
  df        = cover_by_station_taxa,
  table_ref = "uvs.lpi_cover_by_station_taxa",
  label     = "Cover by Station | Taxa",
  filename  = paste0(exp_id, "_uvs_lpi_cover_by_station_taxa.csv")
)
```

## Upload to BigQuery

```{r upload_bq, eval=FALSE}
uploads <- list(
  list(file = stations_csv,     dataset = "uvs", table = "lpi_stations"),
  list(file = observations_csv, dataset = "uvs", table = "lpi_observations"),
  list(file = cover_csv,        dataset = "uvs", table = "lpi_cover_by_station_taxa")
)

for (upload in uploads) {
  table_id <- paste0(upload$dataset, ".", upload$table)
  cat("  Uploading:", table_id, "...")
  
  bq_table_upload(
    bq_table("pristine-seas", upload$dataset, upload$table),
    values = read_csv(upload$file, show_col_types = FALSE),
    create_disposition = "CREATE_NEVER",
    write_disposition  = "WRITE_APPEND"
  )
  
  cat("✓\n")
}
```

---

```{r}
tbl(bq_connection, "uvs.lpi_stations") |> 
  filter(!is.na(exp_id)) |> 
  count(exp_id, sort = T)
```

