---
title: "UVS sites - Data Pipeline"
date: today
format: 
  html:
    theme: minty
    self-contained: true
    code-fold: true
    toc: true 
    toc-depth: 3
    toc-location: right
    html-table-processing: none
execute:
  fig-width: 10
---

```{r setup, message = F, warning = F, fig.width = 10, fig.height = 10, echo = F}
options(scipen = 999)

# Hook to format inline numeric expressions with comma separators:
knitr::knit_hooks$set(inline = function(x) {
  # For non-numeric values, just return as character
  if (!is.numeric(x)) {return(as.character(x))}
  # Format numbers with comma as big.mark, no scientific notation
  else format(x, big.mark = ",", scientific = FALSE)})

library(sf)
library(hms)
library(readxl)
library(janitor)
library(lubridate)
library(gt)
library(pointblank)
library(tidyverse)
library(bigrquery)
library(leaflet)
library(leaflet.extras)
library(ggtext)
library(PristineSeasR2)

# ---- Parameters -----------------------------------------------------------
exp_id   <- "PRT_2018"

ps_paths <- PristineSeasR2::get_drive_paths()

prj_path <- file.path(ps_paths$projects, "legacy-db")

# ---- Auth & Connection ----------------------------------------------------
bigrquery::bq_auth(email = "marine.data.science@ngs.org")
bq_connection <- DBI::dbConnect(bigrquery::bigquery(), project = "pristine-seas")

set_theme(PristineSeasR2::theme_pristineseas())

data_out <- file.path(prj_path, "data/output/uvs")

figures_out <- file.path(prj_path, "figures/uvs")

legacy_path <- file.path(ps_paths$science, "legacy data/Azores/")

exp_config <- list(exp_id      = exp_id,
                   date_bounds = c(as.Date("2018-06-01"), as.Date("2018-06-30")))
```

## Overview

This script processes and validates **Underwater Visual Survey (UVS) site metadata** for the expedition **`r exp_id`**. 

**Workflow**: Import → Standardize → QA/QC → Visualize → Export → Upload

This process ensures data integrity prior to analysis and database ingestion.

## 1. Import and Standardize Fieldbook Data

```{r uvs-import}}
uvs_sites <- read_xlsx(file.path(legacy_path, "FISH", "Azores_fish_data_2018_AMF.xlsx"),
                       sheet = "metadata") |> 
  clean_names() |> 
  rename(latitude  = lat,
         longitude = long,
         subregion = island,
         site      = station) |> 
  filter(!is.na(site))

uvs_sites <- uvs_sites |> 
  mutate(exp_id    = exp_id,
         method    = "uvs",
         date      = lubridate::ymd(date),
         time      = hms::as_hms(time),
         habitat   = str_to_lower(habitat),
         exposure  = str_to_lower(exposure),
         subregion = str_to_title(subregion),
         region    = "Azores",
         locality  = NA_character_,
         in_mpa    = if_else(mangt == "Open", FALSE, TRUE),
         mpa_notes = NA_character_,
         team_lead = "Alan Friedlander",
         site_name = NA_character_,
         ps_site_id =  paste0(exp_id, "_uvs_", str_pad(site, 3, pad = "0"))) |> 
  select(ps_site_id, exp_id, method, region, subregion, locality, 
         date, time, latitude, longitude, habitat, exposure, site_name, in_mpa, mpa_notes, team_lead)

uvs_sites$locality[uvs_sites$ps_site_id %in% c("PRT_2018_uvs_089", "PRT_2018_uvs_091")] <- "Monte da Guia"
uvs_sites$mpa_notes[uvs_sites$ps_site_id %in% c("PRT_2018_uvs_077", "PRT_2018_uvs_079", "PRT_2018_uvs_081")] <- "Paper park"
uvs_sites$habitat <- "rocky_reef"
uvs_sites$exposure <- "unknown"

# From Benthic_Lat_Long_1.xlsx file

uvs_sites$latitude[uvs_sites$ps_site_id == "PRT_2018_uvs_075"] <- 39.02
uvs_sites$longitude[uvs_sites$ps_site_id == "PRT_2018_uvs_075"] <- -29.33

regions_lookup_table <- uvs_sites |> 
  distinct(region, subregion) 

valid_regions <- unique(regions_lookup_table$region)
valid_subregions <- unique(regions_lookup_table$subregion)
```

We conducted a total of `r nrow(uvs_sites)` UVS surveys across `r length(unique(uvs_sites$region))` regions and `r length(unique(uvs_sites$subregion))` subregions. The data includes `r sum(uvs_sites$in_mpa, na.rm = TRUE)` sites within Marine Protected Areas (MPAs). The first site was surveyed on `r min(uvs_sites$date)` and the last on `r max(uvs_sites$date)`.

## 2. QA/QC and Validation

```{r uvs_qaqc}
# 2.QA/QC using pointblank

uvs_sites_qaqc <- uvs_sites |> 
  create_agent(label = "UVS sites QA/QC", tbl_name = "uvs_sites_raw") |> 
  # Ensure each site ID is unique
  rows_distinct(ps_site_id,
                label = "Unique site IDs",
                actions = action_levels(stop_at = 0.001)) |> 
  # Enforce non-missing values for critical columns
  rows_complete(columns = vars(ps_site_id, latitude, longitude, date),
                label = "Complete rows for key fields",
                actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |>
  # Habitat and exposure validation against allowed vocab
  col_vals_in_set(columns = vars(habitat),
                  set = allowed_vocab$uvs_habitats,
                  label = "Valid habitat values",
                  actions = action_levels(stop_at = 0.1)) |>
  col_vals_in_set(columns = vars(exposure),
                  set = allowed_vocab$exposure,
                  label = "Valid exposure values",
                  actions = action_levels(stop_at = 0.1)) |>
  # Region validation against lookup table
  col_vals_in_set(columns = vars(region),
                  set = valid_regions,
                  label = "Region in lookup table",
                  actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |>
  col_vals_in_set(columns = vars(subregion), 
                  set = valid_subregions,
                  label = "Subregion in lookup table",
                  actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |>
  # Date within expedition range
  col_vals_between(columns = vars(date),
                   left = exp_config$date_bounds[1],
                   right = exp_config$date_bounds[2], 
                   label = "Date within expedition period",
                   actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |> 
  # Critical NAs
  col_vals_not_null(columns = vars(latitude, longitude),
                    label = "Coordinates are not missing",
                    actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |>
  col_vals_not_null(columns = vars(date),
                    label = "Date is not missing", 
                    actions = action_levels(warn_at = 0.01, stop_at = 0.05)) |> 
  interrogate()


# Display QA/QC results
uvs_sites_qaqc
```

Validation checks confirmed that all site IDs are unique and coordinates fall within the expedition bounds. 

```{r uvs-missing-data}
#| fig-cap: "Missing data in UVS sites fieldbook"

naniar::vis_miss(uvs_sites)
```

## 3. Visualize

The interactive map below displays unique UVS sites with metadata-rich popups.

```{r uvs-map}
# Interactive visual map

uvs_sites_sf <- uvs_sites |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  distinct(ps_site_id, region, subregion, locality, habitat, exposure, in_mpa, geometry)

mapview::mapview(uvs_sites_sf,
                 zcol = "subregion",
                 legend = TRUE,
                 map.types = "Esri.WorldImagery",
                 layer.name = "subregion",
                 popup = leafpop::popupTable(uvs_sites_sf, 
                                             zcol = c("ps_site_id", "region", "subregion", "locality", "habitat", "exposure", "in_mpa"))) |> 
  leafem::addMouseCoordinates() |> 
  leaflet.extras::addFullscreenControl()
```

## 4. Summary and Export

Of the `r nrow(uvs_sites)` sites, `r sum(uvs_sites$in_mpa, na.rm = TRUE)` are located within MPAs. Surveys covered `r length(unique(uvs_sites$habitat))` habitat types.

```{r}
#| label: tbl-uvs-sites-by-habitat
#| tbl-cap: "Summary of UVS survey effort by region, exposure, and habitat."

uvs_sites |> 
  group_by(region, subregion, habitat) |>
  summarize(n_surveys = n(),
            .groups = "drop") |> 
  pivot_wider(names_from = habitat, values_from = n_surveys, values_fill = 0) |> 
  gt::gt(groupname_col = "leg") |> 
  gt::tab_options(row_group.as_column = T,
                  table.width = pct(80),
                  table.font.size = 12,
                  table.border.top.style = "solid",
                  table.border.top.width = px(1),
                  table.border.top.color = "black",
                  heading.align = "center",
                  column_labels.font.weight = "bold")
```

```{r eval = F}
bq_table_upload(bq_table("pristine-seas", "uvs", "sites"),
                values = uvs_sites,
                create_disposition = "CREATE_NEVER",
                write_disposition  = "WRITE_APPEND")
```

